---
title: "*Faecalibacterium prausnitzii*, depleted in the Parkinson’s disease microbiome, improves motor deficits in alpha-synuclein overexpressing mice"
subtitle: "Data: Metaphlan 4.1.1 relative abundances"
author:
  - name: Anastasiya Moiseyenko
  - name: Aubrey Schonhoff
  - name: Giacomo Antonello
  - name: Kaelyn Long
  - name: Joseph Boktor
  - name: Blake Dirks
  - name: Anastasiya Oguienko
  - name: Alex Villoria-Winnett
  - name: Rustem Ismagilov
  - name: Rosa Krajmalnik-Brown
  - name: Nicola Segata
  - name: Levi Waldron
  - name: Sarkis K. Mazmanian
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::robobook:
    self_contained: true
    toc_depth: 3
    toc_float: false
    use_bookdown: true
    highlight: haddock
params:
  basic_outdir: "results"
  assayTested: "metaphlan_RelAbund"
  microbiome_data_type: "SGB"
  beta_dist: "bray"
  beta_MDS: "MDS"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# packages for tabular data manipulation
library(tidyverse)

# microbiome specific packages
library(maaslin3) # this version needs a custom fix in a fork I made pending approval as of Jul 28 2025. install with remotes::install_github("g-antonello/maaslin3", force = TRUE)
library(mia)
library(biobakeryUtils)
# some packages for data visualization (plots and tables)
library(miaViz)
library(ggpubr)
library(kableExtra)
library(reactable)
library(ggtext) # enhanced text handling in ggplot
# MAC's package to download data
library(parkinsonsMetagenomicData)

library(ggh4x) # for complex ggplot2 faceting
# remotes::install_github("cmartin/ggConvexHull")
library(ggConvexHull) # for convex hulls geometries
# set ggplot2 plot themes 
theme_set(theme_light())
# set seed for reproducibility
set.seed(1234)

library(biobakeryUtils)
###################################################################
# pairwise PERMANOVA source code
source("https://raw.githubusercontent.com/pmartinezarbizu/pairwiseAdonis/refs/heads/master/pairwiseAdonis/R/pairwise.adonis.R")
###################################################################

###################################################################
# function to colorize labels of ggplot object
# This function will apply ALL color rules to EACH label string
colorize_labels <- function(labels, mapping) {
  # Start with the original labels
  colored_labels <- labels
  
  # Loop through each pattern-color rule in your map
  for (pattern in names(mapping)) {
    color <- mapping[[pattern]]
    
    # The text to display is the pattern without the escape slashes
    display_text <- gsub("\\\\", "", pattern) 
    
    # Apply the coloring rule. This will modify labels that have already been partially colored.
    colored_labels <- gsub(
      pattern = pattern, 
      replacement = paste0("<span style='color:", color, ";'>", display_text, "</span>"), 
      x = colored_labels,
      fixed = TRUE
    )
  }
  
  return(colored_labels)
}
###################################################################




overall_color_palette <- c("WT Control" = "#b2df8a", "Thy1-ASO Control" = "#a6cee3", "Thy1-ASO *F. prausnitzii*" = "#1f78b4", "Thy1-ASO F. prausnitzii" = "#1f78b4")

```

# Load data for analysis

```{r}
MetaPhlAn.tse <- read_TSE_from_dir("Data/MetaPhlAn.tse")
```

```{r}
GTDB.df <- read_tsv("Data/GTDB_202403.tsv")
```

# Create output directory

```{r}
dir.create(file.path(params$basic_outdir, params$microbiome_data_type), recursive = TRUE)
```

# Reads distribution per condition

```{r, fig.height=7, fig.width=7}
normality_test <- shapiro.test(MetaPhlAn.tse@colData$number_reads)

hist_reads <- ggplot(colData(MetaPhlAn.tse), aes(x = number_reads/1e6, color = genotype_treatment, fill = genotype_treatment)) + 
  scale_color_manual(values = overall_color_palette, aesthetics = c("fill", "color")) +
  geom_density(alpha = 0.1) + 
  labs(
    x = "Million reads per sample retained",
    y = "Count",
    color = "Genotype and Treatment",
    fill = "Genotype and Treatment",
    caption = sprintf("Shapiro-Wilk: W = %.3f, P = %.3f", normality_test$statistic, normality_test$p.value)
  )

test_reads <- pairwise.t.test(MetaPhlAn.tse@colData$number_reads, MetaPhlAn.tse@colData$genotype_treatment) %>% 
  broom::tidy()

ggarrange(hist_reads, ggtexttable(test_reads, rows = NULL), common.legend = TRUE, labels = "AUTO", nrow = 2, heights = c(1, 0.3))

```

## Alpha diversity does not vary in relation to sequencing depth

### Alpha indices distributions

```{r}
MetaPhlAn.tse <- addAlpha(MetaPhlAn.tse, assay.type = params$assayTested, index = c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"))

lapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), 
       function(div) {
         stats_basic <- shapiro.test(as.data.frame(colData(MetaPhlAn.tse))[[div]])
         text <- paste0("Shapiro-Wilk test: W = ", round(stats_basic$statistic, 2), "; P-value = ", round(stats_basic$p.value, 3))
         plot <- ggplot(as.data.frame(colData(MetaPhlAn.tse)), aes(x = !!sym(div))) +
           geom_histogram() + 
           labs(
             title = div, 
             caption = text
           )
         
         }
       ) %>% 
  ggarrange(plotlist = .)

```

## Statistics

```{r, tab.cap="Assessment of dependence of alpha diversity measures in relation to sequencing depth"}
tmp <- sapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), function(div) {
  lm(as.formula(paste(div, "~", "number_reads")), data = colData(MetaPhlAn.tse))
  
}, USE.NAMES = TRUE, simplify = FALSE)

table_results.df <- tmp %>% lapply(broom::tidy) %>% 
  bind_rows(.id = "AlphaIndex") %>% 
  filter(!grepl("Intercept", term))

table_results.df %>%
  kbl() %>% 
  row_spec(which(table_results.df$p.value < 0.05), bold = TRUE) %>% 
  kable_styling()
```

# Main figure 

## A - Stackplot relative abundances at Phylum Level

```{r}
panelA_stackplot <- plotAbundance(
  x = MetaPhlAn.tse,
  assay.type = params$assayTested, 
  group = "phylum") + 
  facet_grid(cols = vars(genotype_treatment_legend), scales = "free_x") + 
  theme_light() +
  scale_x_discrete(labels = "x") +
  theme(
    panel.spacing.x = unit(0, "lines"),
    strip.background.x = element_rect(fill = "white", colour = "gray40", linewidth = unit(0.8, "mm")), 
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "gray40", linewidth = unit(0.8, units = "mm")),
    strip.text = element_text(color = "black"),
    axis.text.x = element_text(color = "white"),
    axis.ticks.x = element_blank(),
    strip.text.x = element_markdown()
    ) + 
  labs(
    fill = "Phylum",
    x = " "
  )

panelA_stackplot$data$colour_by <- gsub("p__", "", panelA_stackplot$data$colour_by)
panelA_stackplot$data$colour_by <- gsub("_", " ", panelA_stackplot$data$colour_by)
```

## B - PCoA

```{r}
set.seed(1234)
dimRedName <- paste("Fp7", params$beta_dist, "mds", sep = "_")
# this will work once mia or TreeSummarizedExperiment fixes some things

MetaPhlAn.tse <- mia::addMDS(
  MetaPhlAn.tse,
  assay.type = params$assayTested,
  method = params$beta_dist,
  ncomponents = 3, name = dimRedName
  )

MetaPhlAn.tse <- addPERMANOVA(MetaPhlAn.tse, 
             formula = x ~ genotype_treatment, 
             assay.type = params$assayTested,
             method = params$beta_dist,
             permutations = 9999, 
             by = "margin")

variances_explained <- attr(reducedDim(MetaPhlAn.tse, dimRedName), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

Fp7_colData_for_PCoA.df <- cbind.data.frame(as.data.frame(colData(MetaPhlAn.tse)), reducedDim(MetaPhlAn.tse, dimRedName) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

panelB_PCoA <- Fp7_colData_for_PCoA.df %>% 
ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = genotype_treatment_legend, fill = genotype_treatment_legend)) + 
  geom_point(size = 2.5, stroke = 1, color = "black") +
  geom_point(size = 2.5) +
  ggConvexHull::geom_convexhull(show.legend = FALSE, alpha = 0.2) + 
  scale_color_manual(values = overall_color_palette) +
  scale_fill_manual(values = overall_color_palette) +
  theme(legend.position = "top", legend.text = element_markdown()) +
  labs(color = "Genotype and Treatment",
       fill = "Genotype and Treatment",
       x = paste0("PCoA Axis 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("PCoA Axis 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )

# panelB_PCoA
```

## C - Pairwise bray-curtis dissimilarities

### Statistical test (Bray-curtis)

```{r}
dissimName <- paste("Fp7", params$beta_dist, sep = "_")
MetaPhlAn.tse <- addDissimilarity(MetaPhlAn.tse, method = "bray", name = dissimName, assay.type = params$assayTested)
mtx <- metadata(MetaPhlAn.tse)[[dissimName]]
mtx[upper.tri(mtx,diag = TRUE)] <- NA

pairwise_dissim_bray <- reshape2::melt(mtx) %>% 
  filter(complete.cases(.)) %>%
  dplyr::inner_join(., as.data.frame(colData(MetaPhlAn.tse)) %>% transmute(Var1 = uuid, genotype_treatment_legend = genotype_treatment_legend), by = "Var1") %>% 
  dplyr::inner_join(., as.data.frame(colData(MetaPhlAn.tse)) %>% transmute(Var2 = uuid, genotype_treatment_legend = genotype_treatment_legend), by = "Var2") %>% 
  dplyr::select(Var1, Var2, contains(c("uuid", "genotype", "treatment")), value) %>% 
  mutate(pairs = paste(genotype_treatment_legend.y, genotype_treatment_legend.x, sep = " vs ")) %>% 
  filter(genotype_treatment_legend.x != genotype_treatment_legend.y)

```

```{r}
set.seed(1234)
library(vegan)
pairwise_adonis_stats <- pairwise.adonis(x = as.dist(metadata(MetaPhlAn.tse)[[dissimName]]), 
                factors = MetaPhlAn.tse@colData$genotype_treatment, 
                perm = 9999,
                p.adjust.m = "BH"
                )

```

### Boxplot dissimilarities

```{r}
pairwise_adonis_stats_asterisks <- pairwise_adonis_stats %>% 
  mutate(pairs = gsub("Thy1-ASO F. prausnitzii", "Thy1-ASO *F. prausnitzii*", pairs))

panelC_pairwise_dissim_boxplot <- pairwise_dissim_bray %>%
  ggplot(
    aes(x = pairs, y = value)
  ) +
  geom_point(position = position_jitter(width = 0.2, seed = 1234), stroke = 1.1) +
  geom_point(position = position_jitter(width = 0.2, seed = 1234), color = "gray30") +
  geom_boxplot(outliers = FALSE, alpha = 0.6, width = 0.3) +
  geom_text(data = pairwise_adonis_stats_asterisks %>%
              dplyr::left_join(
                group_by(pairwise_dissim_bray, pairs) %>%
                  reframe(value = max(value) + 0.05), by = "pairs") %>% 
              mutate(pairs = factor(pairs, levels = pairs)),
            mapping = aes(x = pairs, y = value, label = sprintf("FDR = %.3f", p.adjusted)), hjust = 0, angle = 90) +
  # Apply the colorize_labels function to your x-axis labels
  scale_x_discrete(labels = function(x) colorize_labels(x, overall_color_palette)) +
  # Tell ggplot to render the axis text as Markdown/HTML
  theme(
    axis.text.x = ggtext::element_markdown(angle = 90, hjust = 1, vjust = 0.5, face = "bold")
    ) +
  ylim(0.3, 0.9) +
  labs(
    x = "",
    y = "Bray-Curtis Pairwise Dissimilarity"
  )

# make a horizontal version just in case

panelC_pairwise_dissim_boxplot_horiz <- pairwise_dissim_bray %>%
  ggplot(
    aes(y = pairs, x = value)
  ) +
  geom_point(position = position_jitter(height = 0.2, seed = 1234), stroke = 1.1) +
  geom_point(position = position_jitter(height = 0.2, seed = 1234), color = "gray30") +
  geom_boxplot(outliers = FALSE, alpha = 0.6, width = 0.3) +
  geom_text(data = pairwise_adonis_stats_asterisks %>%
              dplyr::left_join(
                group_by(pairwise_dissim_bray, pairs) %>%
                  reframe(value = max(value) + 0.05), by = "pairs") %>% 
              mutate(pairs = factor(pairs, levels = pairs)),
            mapping = aes(y = pairs, x = value, label = sprintf("FDR = %.3f", p.adjusted)), hjust = 0) +
  # Apply the colorize_labels function to your x-axis labels
  scale_y_discrete(labels = function(x) colorize_labels(x, overall_color_palette)) +
  # Tell ggplot to render the axis text as Markdown/HTML
  theme(axis.text.y = ggtext::element_markdown(angle = 0, hjust = 1)) +
  xlim(0.3, 0.9) +
  labs(
    y = "",
    x = "Bray-Curtis Pairwise Dissimilarity"
  )

```

## D - Mediation analysis with PCoA axes 1-3 on constipation and motor symptoms

- Constipation symptoms: $Bead~expulsion~(minutes)$
- Motor symptoms: $\frac{slips}{steps}~on~beam$

We use the standard 3-equation approach:

  - (1) $Mediator \sim Treatment$

  - (2) $Outcome \sim Treatment$
  
  - (3) $Outcome \sim Mediator + Treatment$

Where `Mediator` = PCo.x, `Outcome` = Bead Expulsion or Slips/Steps, `Treatment`
*F. prausnitzii* supplementation.

```{r}
# gather bead_expulsion_n data
motor_gi_symptoms.list <- sapply(c("Beam_steps", "Bead_exp"), function(x) openxlsx::read.xlsx("Data/motor_performance/Fp7_motor_performance.xlsx", sheet = x), simplify = FALSE, USE.NAMES = TRUE)

# prepare bead expulsion variable
motor_gi_symptoms.list$Bead_exp <- motor_gi_symptoms.list$Bead_exp %>% 
  dplyr::transmute(
  mouse_id = ID,
  bead_expulsion_min = Trial1)

# prepare slips over steps variable
motor_gi_symptoms.list$Beam_steps <- motor_gi_symptoms.list$Beam_steps %>% 
  rowwise() %>% 
  transmute(
    mouse_id = ID,
    Slips = Slips_Trial1,
    Steps = Steps_Trial1,
    # older choice of presenting data
    # Slips = mean(c(Slips_Trial1, Slips_Trial2), na.rm = TRUE),
    # Steps = mean(c(Steps_Trial1, Steps_Trial2), na.rm = TRUE),
    slips_over_steps = Slips/Steps)

additional_variables.df <- purrr::reduce(motor_gi_symptoms.list, dplyr::full_join, by = "mouse_id")

# incorporate these columns into the colData of the input
new_df <- DataFrame(dplyr::left_join(as.data.frame(colData(MetaPhlAn.tse)), additional_variables.df, by = "mouse_id"))
rownames(new_df) <- new_df$uuid

colData(MetaPhlAn.tse) <- new_df
```

### Mediation analysis of bead expulsion

```{r}
library(mediation)
set.seed(1234)

mediationData.df <- cbind.data.frame(
  as.data.frame(colData(MetaPhlAn.tse)), 
as.data.frame(reducedDim(MetaPhlAn.tse, dimRedName))) %>% 
  dplyr::rename(
    PCo.1 = V1, 
    PCo.2 = V2,
    PCo.3 = V3
    ) %>% 
  filter(
    # remove NAs from covariates of interest
    complete.cases(dplyr::select(., contains("PCo"), genotype_treatment, bead_expulsion_min)), 
    # remove WT Control and use ASO control as baseline
    genotype_treatment != "WT Control"
    ) %>% 
  mutate(
    treatment = droplevels(treatment)
  )

# make 3 mediation analyses as list, one per Principal Component
mediations_Princ.Coord <- sapply(grepv("PCo", colnames(mediationData.df)), function(x) {
  # Model 1 - Effect of exposure on mediator
  mediation.lm <- lm(as.formula(paste(x, "treatment", sep = "~")), data = mediationData.df)
  # Model 2 - Effect of exposure on outcome (Total effect)
  outcome.lm <- lm(as.formula(paste("bead_expulsion_min ~ treatment",  x, sep = "+")), data = mediationData.df)
  # Model 3 - Meidation analysis
  mediation_analysis = mediate(
        model.m = mediation.lm,
        model.y = outcome.lm,
        sims = 1000,
        treat = "treatment",
        mediator = x,
        boot = FALSE,
        robustSE = TRUE
      )
  # return all 3 models for completeness
  
  return(
    list(
      mediator.lm = mediation.lm,
      outcome.lm = outcome.lm,
      mediation_analysis = mediation_analysis
    )
  )
}, simplify = FALSE, USE.NAMES=TRUE)

data_as_long_table <- mediations_Princ.Coord %>% 
  lapply("[[", "mediation_analysis") %>% 
  lapply(tidy_mediation) %>% 
  bind_rows(.id = "PCo")

write_tsv(data_as_long_table, file = file.path(params$basic_outdir, params$microbiome_data_type, "PCo1-3_mediation_bead_expulsion_min_ASO_treat_vs_ASO_ctrl.tsv"))
```

### Mediation analysis of slips/steps

```{r}
library(mediation)
set.seed(1234)

mediationData.df <- cbind.data.frame(
  as.data.frame(colData(MetaPhlAn.tse)), 
as.data.frame(reducedDim(MetaPhlAn.tse, dimRedName))) %>% 
  dplyr::rename(
    PCo.1 = V1, 
    PCo.2 = V2,
    PCo.3 = V3
    ) %>% 
  filter(
    # remove NAs from covariates of interest
    complete.cases(dplyr::select(., contains("PCo"), genotype_treatment, slips_over_steps)), 
    # remove WT Control and use ASO control as baseline
    genotype_treatment != "WT Control"
    ) %>% 
  mutate(
    treatment = droplevels(treatment)
  )

# make 3 mediation analyses as list, one per Principal Component
mediations_Princ.Coord <- sapply(grepv("PCo", colnames(mediationData.df)), function(x) {
  # Model 1 - Effect of exposure on mediator
  mediation.lm <- lm(as.formula(paste(x, "treatment", sep = "~")), data = mediationData.df)
  # Model 2 - Effect of exposure on outcome (Total effect)
  outcome.lm <- lm(as.formula(paste("slips_over_steps ~ treatment",  x, sep = "+")), data = mediationData.df)
  # Model 3 - Meidation analysis
  mediation_analysis = mediate(
        model.m = mediation.lm,
        model.y = outcome.lm,
        sims = 1000,
        treat = "treatment",
        mediator = x,
        boot = FALSE,
        robustSE = TRUE
      )
  # return all 3 models for completeness
  
  return(
    list(
      mediator.lm = mediation.lm,
      outcome.lm = outcome.lm,
      mediation_analysis = mediation_analysis
    )
  )
}, simplify = FALSE, USE.NAMES=TRUE)

data_as_long_table <- mediations_Princ.Coord %>% 
  lapply("[[", "mediation_analysis") %>% 
  lapply(tidy_mediation) %>% 
  bind_rows(.id = "PCo") 

write_tsv(data_as_long_table, file = file.path(params$basic_outdir, params$microbiome_data_type, "PCo1-3_mediation_slips_over_steps_ASO_treat_vs_ASO_ctrl.tsv"))
```

### Generate plot

```{r}
data_as_long_table_slips_over_steps <- read_tsv(file.path(params$basic_outdir, params$microbiome_data_type, "PCo1-3_mediation_slips_over_steps_ASO_treat_vs_ASO_ctrl.tsv")) %>% 
  mutate(Exposure = "Slips/Steps Ratio")

data_as_long_table_bead_expulsion <- read_tsv(file.path(params$basic_outdir, params$microbiome_data_type, "PCo1-3_mediation_bead_expulsion_min_ASO_treat_vs_ASO_ctrl.tsv")) %>% 
  mutate(Exposure = "Bead Expulsion (minutes)")


data_as_long_table_for_horizPlot <- rbind.data.frame(data_as_long_table_slips_over_steps,data_as_long_table_bead_expulsion) %>% 
  mutate(
    PCo = factor(PCo, levels = sort(unique(data_as_long_table$PCo), decreasing = TRUE)),
    Effect = factor(Effect, levels = sort(unique(Effect), decreasing = TRUE))
         )
```

```{r}
panelD_mediation_motor_GI_symptoms <- ggplot(data_as_long_table_for_horizPlot, aes(x = Estimate, y = Effect, color = PCo, shape = P.value < 0.05)) + 
  geom_vline(xintercept = 0, lty = "dashed", color = "darkgray") + 
  geom_errorbar(aes(xmin = CI_Lower_95, xmax = CI_Upper_95, y = Effect, color = PCo), position = position_dodge(width = 0.3, preserve = "total"), width = 0, show.legend = FALSE) + 
  geom_point(position = position_dodge(width = 0.3, preserve = "total"), size = 4) + 
  scale_shape_manual(values = c(1,19)) +
  facet_nested(cols = vars(Exposure), scales = "free_x") +
  scale_color_manual(values = c("darksalmon", "brown3", "brown4"), labels = c("PCo 3", "PCo 2", "PCo 1")) +
  guides(color = guide_legend(reverse = TRUE), shape = "none") +
  theme(
    panel.spacing.x = unit(0, "lines"),
    strip.background.x = element_rect(fill = "white", colour = "gray40", linewidth = unit(0.8, "mm")), 
    strip.text = element_text(color = "black"), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12)
    ) +
  labs(
    color = NULL,
    y = NULL,
    x = expression(beta~Estimate)
    )

panelD_mediation_motor_GI_symptoms

panelD_mediation_motor_GI_symptoms_shorterLabels <- panelD_mediation_motor_GI_symptoms + scale_y_discrete(labels = function(x) gsub("Average\\ |Causal\\ ", "", x))
  

```

## E - Differentially abundant features due to treatment in Thy1-ASO mice

Model formula is:

$$
\log_2(Microbiome)_{Thy1-ASO} \sim treament  
$$

```{r}
# prepare input data
tmp.tse <- MetaPhlAn.tse[,MetaPhlAn.tse@colData$genotype != "WT"]
tmp.tse@colData$treatment <- relevel(tmp.tse@colData$treatment, "Control")
tmp.tse@colData$treatment <- droplevels(tmp.tse@colData$treatment)
tmp.tse <- tmp.tse[rowSums(assay(tmp.tse, params$assayTested)) > 0,]

tmp.tse

# define experiment name and output directories
maaslin3_expmt <- "treatments_vs_Control_in_Thy1-ASO"
maaslin3_final_outdir <- file.path(params$basic_outdir, params$microbiome_data_type, maaslin3_expmt)

# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
set.seed(1234)
treatments_vs_Control_in_Thy1ASO.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ treatment, 
    transform = "LOG",
    output = maaslin3_final_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE, 
    plot_associations = TRUE,
    save_plots_rds = TRUE,
		verbosity = "ERROR",
		summary_plot_first_n = 15
  )

## add taxonomy to output tables: all_results.tsv
dplyr::right_join(GTDB.df %>% mutate(SGB = paste0("t__", SGB)), read_tsv(file.path(maaslin3_final_outdir, "all_results.tsv")) %>% dplyr::rename(!!sym(params$microbiome_data_type) := feature), by = params$microbiome_data_type) %>% 
  #arrange again based on increasing qval_individual
  arrange(qval_individual) %>% 
  write_tsv(file.path(maaslin3_final_outdir, "all_results.tsv"))

## add taxonomy to output tables: significant_results.tsv
dplyr::right_join(GTDB.df %>% mutate(SGB = paste0("t__", SGB)), read_tsv(file.path(maaslin3_final_outdir, "significant_results.tsv")) %>% dplyr::rename(!!sym(params$microbiome_data_type) := feature), by = params$microbiome_data_type) %>% 
  write_tsv(file.path(maaslin3_final_outdir, "significant_results.tsv"))

# save raw output
saveRDS(treatments_vs_Control_in_Thy1ASO.maaslin3, file = file.path(maaslin3_final_outdir, "maaslin3_raw_output.rds"))
```

```{r}
GTDB.df <- GTDB.df %>%
  mutate(
    altNames = sapply(
      strsplit(paste(trimws(Genus), trimws(Species), sprintf("[%s]", SGB)), " ", fixed = TRUE),
      function(x) paste(unique(x), collapse = " ")
    )
  )

altNames.chr <- GTDB.df$altNames
names(altNames.chr) <- GTDB.df$SGB

panelE_diffAbund <- readRDS(file = file.path(maaslin3_final_outdir, "figures", "summary_plot_gg.RDS")) + 
  theme(strip.text = element_blank()) + 
  # expand plot to show full errorbars
  scale_x_continuous(expand = expansion(mult = c(.1))) + 
  # remove the null HP that would mess with the panel rendering
  guides(linetype = "none")



panelE_diffAbund2 <- panelE_diffAbund +
  scale_y_discrete(
    labels = function(y) altNames.chr[gsub("t__", "", y)]
  ) + theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )
```

## Main Figure Panel

```{r, fig.height=8, fig.width=12}

panel_A_legend <- get_legend(panelA_stackplot + guides(fill = guide_legend(nrow = 2, title = "Phylum", position = "top")))

Microbiome_Main_Figure <- ggarrange(

  ggarrange(
    panelA_stackplot + guides(fill = guide_legend(title = "Phylum")) + labs(y = "Relative Abundance", x = "Mouse") + theme(axis.title = element_text(size = 12)), 
    panelB_PCoA + theme(axis.title = element_text(size = 12)), 
    common.legend = TRUE, widths = c(1, 1), legend = "top", labels = c("A", "B")),
    
    ggarrange(
      panelC_pairwise_dissim_boxplot_horiz  + theme(axis.title = element_text(size = 12), axis.text = element_text(size = 9)),
      ggarrange(
        get_legend(panelB_PCoA + guides(color = guide_legend(ncol = 1, position = "right"))),
        ggplot() + theme_void(), nrow = 2, heights = c(1, 0.35)
        ), 
      
      nrow = 1, widths = c(1,0.2), labels = c("C", "")
      ),
  
  ggarrange(
  ggarrange(
    panelD_mediation_motor_GI_symptoms_shorterLabels + theme(legend.position = "bottom"),
    ggplot() + theme_void(),
    panelE_diffAbund2 + 
    coord_flip() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), axis.title = element_text(hjust = 1),
          legend.position = "none") + 
    labs(x = expression(log[2]~Fold~Change), y = NULL) + 
    theme(axis.title = element_text(size = 12)), 
    widths = c(1, 0.05, 0.85),
    labels = c("D", "E",  ""), nrow = 1
    ),
  get_legend(panelE_diffAbund2 + theme(legend.position = "bottom", legend.text = element_text(size = 10), legend.title = element_text(size = 12))), nrow = 2,
  heights = c(1, 0.2)
  ),
  
  nrow = 3, 
  heights = c(1, 0.25, 1)
  )

ggsave(plot = Microbiome_Main_Figure, filename = file.path(params$basic_outdir, params$microbiome_data_type, "Microbiome_Main_Figure.svg"), width = 11, height = 13)

```

# Supplementary Figure Panel

```{r}
alpha_data.df <- MetaPhlAn.tse %>%
  colData() %>%
  as.data.frame() %>% 
  dplyr::rename(
    `Shannon Diversity` = shannon_diversity,
    `Observed Richness` = observed_richness
  )

comparisons <- combn(levels(alpha_data.df$genotype_treatment_legend), 2, simplify = FALSE)
```

## A - Observed Richness

```{r}
stats_obs <- rstatix::pairwise_wilcox_test(alpha_data.df, formula = `Observed Richness` ~ genotype_treatment,p.adjust.method =  "BH") %>% 
  mutate(group2 = gsub("F. prausnitzii", "*F. prausnitzii*", group2),
         p.adj.round3 = round(p.adj, 3)) %>% 
  rstatix::add_y_position(step.increase = 0.5)

anova_test <- broom::tidy(aov(`Observed Richness` ~ genotype_treatment, data = alpha_data.df))

panelA_Obs_boxplot <- alpha_data.df %>% 
  ggplot(aes(x = genotype_treatment_legend, y = `Observed Richness`, color = genotype_treatment_legend, fill = genotype_treatment_legend)) + 
  geom_boxplot(outliers = FALSE, show.legend = FALSE, color = "black", alpha = 0.4) + 
  geom_point(color = "black", stroke = 1.1, position= position_jitterdodge(jitter.width = 0.5, seed = 1234)) + 
  geom_point(position= position_jitterdodge(jitter.width = 0.5, seed = 1234)) + 
  scale_color_manual(values = overall_color_palette) +
  scale_fill_manual(values = overall_color_palette) +
  # stat_pvalue_manual(stats_obs, label = "p.adj.round3", tip.length = 0.01) + 
  theme(axis.text.x = element_blank(), legend.text = element_markdown()) +
  labs(
    x = NULL,
    color = "Genotype and Treatment",
    fill = "Genotype and Treatment",
    caption = sprintf("ANOVA F =  %.3f, P-value = %.3f", anova_test$statistic, anova_test$p.value)
  )
```

## B - Shannon Diversity

```{r}
stats_shannon <- rstatix::pairwise_wilcox_test(alpha_data.df, formula = `Shannon Diversity` ~ genotype_treatment, p.adjust.method =  "BH") %>% 
  mutate(group2 = gsub("F. prausnitzii", "*F. prausnitzii*", group2),
         p.adj.round3 = round(p.adj, 3)) %>% 
  rstatix::add_y_position(step.increase = 0.5)

anova_test <- broom::tidy(aov(`Shannon Diversity` ~ genotype_treatment, data = alpha_data.df))

panelB_Shannon_boxplot <- alpha_data.df %>% 
  ggplot(aes(x = genotype_treatment_legend, y = `Shannon Diversity`, color = genotype_treatment_legend, fill = genotype_treatment_legend)) + 
  geom_boxplot(outliers = FALSE, show.legend = FALSE, color = "black", alpha = 0.4) + 
  geom_point(color = "black", stroke = 1.1, position= position_jitterdodge(jitter.width = 0.5, seed = 1234)) + 
  geom_point(position= position_jitterdodge(jitter.width = 0.5, seed = 1234)) + 
  scale_color_manual(values = overall_color_palette) +
  scale_fill_manual(values = overall_color_palette) +
  # stat_pvalue_manual(stats_shannon, label = "p.adj.round3", tip.length = 0.01) + 
  theme(axis.text.x = element_blank(), legend.text = element_markdown()) +
  labs(
    x = NULL,
    color = "Genotype and Treatment",
    fill = "Genotype and Treatment",
    caption = sprintf("ANOVA F =  %.3f, P-value = %.3f", anova_test$statistic, anova_test$p.value)
  )
```

## C - Firmicutes/Bacteroidetes Ratio

```{r}
assayextractedPhylum <- assay(agglomerateByRank(MetaPhlAn.tse, "phylum"), params$assayTested)
MetaPhlAn.tse$FB.Ratio <- assayextractedPhylum["p__Firmicutes",]/assayextractedPhylum["p__Bacteroidota",]

anova_test <- broom::tidy(aov(FB.Ratio ~ genotype_treatment, data = colData(MetaPhlAn.tse)))

panelF_FB.Ratio.df <- colData(MetaPhlAn.tse) %>% 
  as.data.frame() %>% 
  dplyr::select(genotype_treatment, genotype_treatment_legend, FB.Ratio)

stats_fbrartio <- rstatix::pairwise_wilcox_test(panelF_FB.Ratio.df, formula = FB.Ratio ~ genotype_treatment, p.adjust.method =  "BH") %>% 
  mutate(group2 = gsub("F. prausnitzii", "*F. prausnitzii*", group2),
         p.adj.round3 = round(p.adj, 3)) %>% 
  rstatix::add_y_position(step.increase = 0.5)


panelC_FB.Ratio <- panelF_FB.Ratio.df %>% 
  ggplot(aes(x = genotype_treatment_legend, y = FB.Ratio, color = genotype_treatment_legend, fill = genotype_treatment_legend)) + 
  geom_boxplot(outliers = FALSE, show.legend = FALSE, color = "black", alpha = 0.4) + 
  geom_point(color = "black", stroke = 1.1, position= position_jitterdodge(jitter.width = 0.5, seed = 1234)) + 
  geom_point(position= position_jitterdodge(jitter.width = 0.5, seed = 1234)) + 
  scale_color_manual(values = overall_color_palette) +
  scale_fill_manual(values = overall_color_palette) +
  # stat_pvalue_manual(stats_fbrartio, label = "p.adj.round3", tip.length = 0.01) + 
  theme(axis.text.x = element_blank(), legend.text = element_markdown()) +
  labs(
    x = NULL,
    y = "Firmicutes/Bacteroidetes Ratio",
    color = "Genotype and Treatment",
    fill = "Genotype and Treatment",
    caption = sprintf("ANOVA F =  %.3f, P-value = %.3f", anova_test$statistic, anova_test$p.value)
  )

```


## Supplementary Figure Panel

```{r}
microbiome_panel_Supplementary <- ggarrange(
    panelA_Obs_boxplot + theme(axis.ticks.x = element_blank()) + scale_y_continuous(expand = expansion(mult = 0.08)),
    panelB_Shannon_boxplot + theme(axis.ticks.x = element_blank()) + scale_y_continuous(expand = expansion(mult = 0.08)),
    panelC_FB.Ratio + theme(axis.ticks.x = element_blank()) + scale_y_continuous(expand = expansion(mult = 0.08)),
    nrow = 1, labels = c("A", "B", "C"), 
common.legend = TRUE)

ggsave(plot = microbiome_panel_Supplementary, filename = file.path(params$basic_outdir, params$microbiome_data_type, "Supplementary Figure 1 - alpha diversities and FB Ratio.svg"), width = 10, height = 4)

microbiome_panel_Supplementary
```

# Supplementary Tables

```{r}
tableIndex <- 1
```

## Supplementary Table `r tableIndex` - Genome Taxonomy Database of the SGBs in the study

```{r}
write_tsv(GTDB.df[match(gsub("t__", "", rownames(MetaPhlAn.tse)), GTDB.df$SGB),], file = file.path(params$basic_outdir, params$microbiome_data_type, sprintf("Supplementary Table %i - GTDB taxonomy of the study.tsv", tableIndex)))

tableIndex <- tableIndex + 1
```

## Supplementary Table `r tableIndex` - PERMANOVA relative to Main Figure Panel B

```{r}
permanova_table.df <- PERMANOVA_to_table(MetaPhlAn.tse)

write_tsv(permanova_table.df, file = file.path(params$basic_outdir, params$microbiome_data_type, sprintf("Supplementary Table %i - PERMANOVA of %s distance.tsv", tableIndex, params$beta_dist)))

tableIndex <- tableIndex  + 1
```

## Supplementary table `r tableIndex` - Pairwise PERMANOVA statistics

```{r}
write_tsv(pairwise_adonis_stats, file = file.path(params$basic_outdir, params$microbiome_data_type, sprintf("Supplementary Table %i - pairwise PERMANOVA of %s distance.tsv", tableIndex, params$beta_dist)))

tableIndex <- tableIndex  + 1
```

# Session Info

```{r}
sessionInfo()
```

  