---
title: "Probiotics to the rescue: F. prausnitzii and beneficial communities to re-establish motor performance in Parkinson mouse models"
author: "Anastasia Moiseyenko, Giacomo Antonello, Nicola Segata, Levi Waldron, Sarkis Mazmanian"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
params:
  basic_outdir: "results"
  microbiome_basic_transform: "RelAbund"
  tax_level: "SGB"
  beta_dist: "bray"
  beta_MDS: "MDS"
---

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
# packages for tabular data manipulation
library(tidyverse)
library(data.table)

# microbiome specific packages
library(maaslin3)
library(mia)
# some packages for data visualization (plots and tables)
library(miaViz)
library(pheatmap)
library(UpSetR)
library(ggpubr)
library(kableExtra)
library(reactable)

# MAC's package to download data
library(parkinsonsMetagenomicData)

library(ggh4x) # for complex ggplot2 faceting
# set ggplot2 plot themes 
theme_set(theme_light())
# set seed for reproducibility
set.seed(1234)
```

# Create output directories all in one chunk

```{r}

output_directories.list <- list(
  # alpha diversity directory
  alpha_div_outdir = file.path(params$basic_outdir, params$tax_level, "01_alpha_diversity"),
  # beta diversity output directory
  beta_div_outdir = file.path(params$basic_outdir, params$tax_level, "02_beta_diversity"),
  # differential abundance temporary directory
  maaslin_tmpdir = "~/Documents/maaslin3_tmp_results/mazmanian_Ana"
)

tmp_out <- capture.output(sapply(output_directories.list, dir.create, showWarnings = FALSE, recursive = TRUE))
rm(tmp_out)
```

# Prepare data for analysis

THis is run only if interactive to avoid careless overwriting

```{r, eval=interactive()}

# Load metadata
metadata.df <- read_tsv("Data/metadata_4.0.tsv") %>% 
  # format a few values in the metadata
  mutate(
    genotype = as.factor(genotype) %>% relevel(ref = "WT"),
    cage = as.factor(cage),
    cohort = as.factor(cohort),
    subcohort = as.factor(subcohort),
    treatment = as.factor(treatment) %>% relevel(ref = "Vehicle")
    )

googleCloudStorageR::gcs_auth(json_file = "~/Downloads/curatedmetagenomicdata-232f4a306d1d.json")

# METAPHLAN4
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/get_reads_processed.R")
mpa_files_cache <- cacheMetagenomicData(metadata.df$uuid, data_type = "relative_abundance", ask_on_update = FALSE) %>% 
  # retrieve reads per sample too
  mutate(n_reads_processed = get_reads_processed.vec(cache_path))

# metaphlan <- get_metaphlan_locators(metadata.df$uuid, data_type = "relative_abundance")
metaphlan_profiles <- suppressMessages(loadMetagenomicData(mpa_files_cache))

# the order of samples is preserved this way
metadata.df <- as.data.frame(dplyr::left_join(metadata.df, select(mpa_files_cache, uuid = UUID, n_reads_processed), by = "uuid"))
rownames(metadata.df) <- metadata.df$uuid
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/wrangle_metaphlan.R")
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/complete_unknown_taxonomy.R")

mpa_raw <- cbind(
  as.data.frame(rowData(metaphlan_profiles)) %>%
    rownames_to_column("clade_name"), assay(metaphlan_profiles)) %>% 
  as_tibble() %>% 
  select(-additional_species)

dir.create("Data/metaphlan_profiles", recursive = TRUE, showWarnings = FALSE)
write_tsv(mpa_raw, "Data/metaphlan_profiles/metaphlan4_data.tsv")

# wrangle metaphlan4 data
mpa4_wrangled_taxonomy_mpa <- assay(metaphlan_profiles) %>%
  as.data.frame() %>% 
  rownames_to_column("clade_name") %>% 
  wrangle_metaphlan(taxonomic_lvl = params$tax_level)

InputData.tse <- TreeSummarizedExperiment(
  rowData = mpa4_wrangled_taxonomy_mpa$taxonomies,
  colData = metadata.df,
  assays = list(
    metaphlan_RelAbund = as.matrix(mpa4_wrangled_taxonomy_mpa$profiles/100)[,rownames(metadata.df)],
    metaphlan_Percent = as.matrix(mpa4_wrangled_taxonomy_mpa$profiles)[,rownames(metadata.df)]
    )
)

# Add phylogenetic tree
library(TreeTools)
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/AddPhyloTree_to_mpa_tse.R")
InputData.tse <-  AddPhyloTree_to_mpa_tse(InputData.tse, CHOCOPhlAn_version = "202403")

detach("package:TreeTools", unload=TRUE)
# create a new assay with approximated counts as relAbund * reads_processed

identical(rownames(InputData.tse@colData), colnames(assay(InputData.tse, "metaphlan_RelAbund")))
assay(InputData.tse, "metaphlan_Counts") <- t(round(apply(assay(InputData.tse, "metaphlan_RelAbund"), 1, function(x) x * InputData.tse@colData$n_reads_processed), 0))
saveRDS(InputData.tse, "Data/InputData.tse.Rds")

```

```{r}
InputData.tse <- readRDS("Data/InputData.tse.Rds")
InputData.tse
```

# Dataset-wide microbiome composition statistics

## Number of taxa per taxonomic level

```{r}
InputData_splitByRanks.tse <- append(splitByRanks(InputData.tse), list("SGB" = InputData.tse))
 
InputData_splitByRanks.tse %>%  
  sapply(nrow)
```

## Prevalence distribution of each taxon

```{r}
InputData_splitByRanks.tse %>% 
  lapply(assay) %>% 
  lapply(apply, 1, function(x) sum(x > 0)) %>% 
  sapply(summary)
```

## Relative abundance of 30 most abundant bacteria per taxonomic level {.tabset}

```{r}
tmp <- InputData_splitByRanks.tse %>% 
  lapply(function(tse){
    max_n <- nrow(tse)
    top30MostAbund <- getTop(tse, assay.type = "metaphlan_RelAbund", method = "mean", top = min(c(30, max_n)))
    tse.subset <- tse[top30MostAbund,]
    return(tse.subset)
  })
  
```

```{r, results = "asis"}

for (tx in names(tmp)) {
  # Create header with cat()
  cat("\n\n###", tx, "\n")
  
  # Create plot and convert to plotly
  p <- miaViz::plotAbundance(tmp[[tx]], 
                            assay.type = "metaphlan_RelAbund",
                            group = tx) +
       facet_grid(cols = vars(cohort, treatment), scales = "free_x") + 
    theme(legend.position = "bottom")
  
  # Explicitly print plotly object
  print(p)
  
  cat("\n")
}

```

This is again to show that there are too many taxa to easily visualize them.
We need to rely on other methods.

## Heatmap all taxa and samples

```{r, fig.cap="Heatmap of full microbiome composition of the dataset. Relative abundances have been Z-scaled by taxon, to account for intrinsic differences in relative abundances in the microbiome"}

pheatmap::pheatmap(assay(InputData.tse) %>% magrittr::set_colnames(colData(InputData.tse)$sample_name), scale = "row", annotation_col = colData(InputData.tse) %>% as.data.frame() %>%  magrittr::set_rownames(.$sample_name) %>% select(treatment, cohort), angle_col = 315, show_rownames = FALSE)

```

# Analyses were be performed at the `r params$tax_level` level

Below the `TreeSummarizedExperiment` object summary that is going to be used
in all following analyses

```{r}
InputData.tse <- InputData_splitByRanks.tse[[params$tax_level]]

InputData.tse
```

# Alpha diversity does not vary in relation to sequencing depth

```{r}
InputData.tse <- addAlpha(InputData.tse, assay.type = "metaphlan_RelAbund", index = c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"))

lapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), 
       function(div) {
         stats_basic <- shapiro.test(as.data.frame(colData(InputData.tse))[[div]])
         text <- paste0("Shapiro-Wilk test for normality: W = ", round(stats_basic$statistic, 2), "; P-value = ", round(stats_basic$p.value, 3))
         plot <- ggplot(as.data.frame(colData(InputData.tse)), aes(x = !!sym(div))) +
           geom_histogram() + 
           annotate(geom = "text", x = -Inf, y = Inf, label = text, hjust = 0.2, vjust = 2) + 
           labs(title = div)
         
         }
       ) %>% 
  ggarrange(plotlist = .)

```

```{r, tab.cap="Assessment of dependence of alpha diversity measures in relation to sequencing depth"}
tmp <- sapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), function(div) {
  lm(as.formula(paste(div, "~", "n_reads_processed")), data = colData(InputData.tse))
  
}, USE.NAMES = TRUE, simplify = FALSE)

table_results.df <- tmp %>% lapply(broom::tidy) %>% 
  bind_rows(.id = "AlphaIndex") %>% 
  filter(!grepl("Intercept", term))

table_results.df %>%
  kbl() %>% 
  row_spec(which(table_results.df$p.value < 0.05), bold = TRUE) %>% 
  kable_styling()
```

# Alpha diversity does not vary in relation to genotype, treatment and cohort

$$
AlphaIndex \sim genotype + treatment + cohort + \text{n_reads_processed}
$$

```{r}
tmp <- sapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), function(div) {
  lm(as.formula(paste(div, "~", "genotype + treatment + cohort + n_reads_processed")), data = colData(InputData.tse))
  
}, USE.NAMES = TRUE, simplify = FALSE)

table_results.df <- tmp %>% lapply(broom::tidy) %>% 
  bind_rows(.id = "AlphaIndex") %>% 
  filter(!grepl("Intercept", term)) %>% 
  arrange(term) %>% 
  relocate(term)

write_tsv(table_results.df, file = file.path(output_directories.list$alpha_div_outdir, "alphas_vs_genotype_cohort_treatment.tsv"))

table_results.df %>%
  kbl() %>% 
  row_spec(which(table_results.df$p.value < 0.05), bold = TRUE, color = "red3") %>% 
  kable_styling()
```


# Variables and potential confounders

Variables:

  * treatment: the probiotic community given, or the empty liquid vehicle gavaged
  * genotype: mice genotype, either WT or ASO

Possible confounders:
  
  * Cohort: important to consider because it could determine inter-testing variability
  * Cage: Potential confounder, but mice receiving the same treatment are also in the same cage and there are only 2 mice per cage, making statics weak
  * Subcohort: Same as cage. It looks like a few samples were processed at a different time: 7W-F, 8W-F, 8W-B, 8C-B

# Beta diversity in relation to treatment and cohort

```{r}
dimRedName <- paste(params$beta_dist, "mds", sep = "_")
# this will work once mia or TreeSummarizedExperiment fixes some things
InputData.tse <- mia::addMDS(
  InputData.tse,
  assay.type = "metaphlan_RelAbund",
  method = params$beta_dist,
  ncomponents = 3, name = dimRedName
  )

variances_explained <- attr(reducedDim(InputData.tse, dimRedName), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

project1_meta_with_MDS.df <- cbind.data.frame(as.data.frame(colData(InputData.tse)), reducedDim(InputData.tse, dimRedName) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

beta_plot_basic <- project1_meta_with_MDS.df %>% 
ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = paste(genotype, treatment))) + 
  geom_point(size = 2) + 
  stat_ellipse(show.legend = FALSE) + 
  scale_color_manual(values = ggsci::pal_jco()(5)[c(3, 1, 2, 4, 5)]) +
  theme(legend.position = "top") +
  labs(color = "Genotype and Treatment",
       x = paste0("MDS 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("MDS 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )

```

## Static plot

```{r}
ggsave(plot = beta_plot_basic, filename = file.path(output_directories.list$beta_div_outdir, "bray_MDS_plot.png"), dpi = 600, height = 6.5, width = 7)

beta_plot_basic
```

There are not major microbiota shifts due to probiotic treatment. This is 
expected, as bidimensional separation can usually be achieved in extreme 
treatments. Interestingly, ASO vehicle and WT vehicle show very similar 
compositions.

Matheus' data shows a slightly clearer shift (but still not a separation)

## PERMANOVA statistics

```{r}
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/PERMANOVA_to_table.R")

InputData.tse <- addPERMANOVA(
  InputData.tse,
  assay.type = "metaphlan_RelAbund",
  method = "bray",
  formula = x ~ genotype + treatment + cohort,
  name = "permanova",
  permutations = 999,
  test.homogeneity = TRUE
)

permanova_table.df <- PERMANOVA_to_table(InputData.tse)

write_tsv(permanova_table.df, file = file.path(output_directories.list$beta_div_outdir, paste(params$beta_dist, "adonisPermanova", "genotype", "treatment", "cohort.tsv", sep = "_")))
```

```{r}
permanova_table.df %>% 
  kbl() %>% 
  kable_styling()
```

Genotype and cage show similar microbiome variance explained (R2). The treatment
shows 2x more variance explained. Each variable is tested against the remaining
variables in the model.

# What microbes differ between `genotypes` in vehicle controls?

These bacteria should not be considered in the main exposure of interest later, 
as they vary because of genotype as well. 

Model formula is:

$$
\log_2(Microbiome) \sim genotype
$$

```{r}
# prepare input data
tmp.tse <- InputData.tse[,InputData.tse@colData$treatment == "Vehicle"] 
tmp.tse@colData$genotype <- relevel(tmp.tse@colData$genotype, "WT")
tmp.tse@colData$genotype <- droplevels(tmp.tse@colData$genotype)

# define experiment name and output directories
maaslin3_expmt <- "genotype_differences_in_ctrls"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, maaslin3_expmt)
# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)
# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
genotype_differences_in_ctrls.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ genotype + cohort, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE, 
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

genotype_differences_in_ctrls.maaslin3_annotated <- list(
      abundance = genotype_differences_in_ctrls.maaslin3$fit_data_abundance$results %>% filter(metadata == "genotype"),
      prevalence = genotype_differences_in_ctrls.maaslin3$fit_data_prevalence$results %>% filter(metadata == "genotype")
    ) %>% 
      lapply(function(x) {
    taxonomy_info <- as.data.frame(rowData(tmp.tse))
    taxonomy_info_subset <- taxonomy_info[,1:which(colnames(taxonomy_info) == params$tax_level)]
    colnames(taxonomy_info_subset)[ncol(taxonomy_info_subset)] <- "feature"
    
    return(
      dplyr::full_join(taxonomy_info_subset, x, by = "feature") %>%
        arrange(qval_individual))}
    )

# save output tsvs as produced by maaslin3
system(paste0("cp ", maaslin3_tmp_outdir, "/*.tsv", " ", file.path(params$basic_outdir, params$tax_level, maaslin3_expmt)))
# save figures folder
system(paste0("cp -r ", file.path(output_directories.list$maaslin_tmpdir, maaslin3_expmt, "/figures"), " ", file.path(params$basic_outdir, params$tax_level, maaslin3_expmt)))

```

## Disappeared bacteria in ASO 

```{r}
genotype_differences_in_ctrls.maaslin3_annotated$prevalence %>% 
  filter(qval_individual < 0.1, value == "ASO", round(coef,4) == min(round(coef, 4), na.rm = TRUE)) %>% 
  select(Phylum:feature, coef, qval_individual) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```

## Appeared bacteria in ASO 

```{r}
genotype_differences_in_ctrls.maaslin3_annotated$prevalence %>% 
  filter(qval_individual < 0.1, value == "ASO", round(coef,4) == max(round(coef, 4), na.rm = TRUE)) %>% 
  select(Phylum:feature, coef, qval_individual) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```

## Decreased bacteria in ASO

```{r}
genotype_differences_in_ctrls.maaslin3_annotated$abundance %>% 
  filter(qval_individual < 0.1, value == "ASO", coef < 0) %>% 
  select(Phylum:feature, coef, qval_individual) %>%
  arrange(desc(abs(coef))) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```

## Increased bacteria in ASO

```{r}
genotype_differences_in_ctrls.maaslin3_annotated$abundance %>% 
  filter(qval_individual < 0.1, value == "ASO", coef > 0) %>% 
  select(Phylum:feature, coef, qval_individual) %>%
  arrange(desc(abs(coef))) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```

## Conclusions

  * With the formula $Microbiome \sim genotype + cohort$ we do not see 
  statistically associated bacteria
  
  * Nor having only genotype as exposure (Not shown)
  
  * WT and ASO mice in this experiment do not show statistically significant 
  microbiome differences (N = 16 vs 16)
  
# Effects of `treatment` adjusting for cohort in ASO male mice

Model formula is:

$$
\log_2(Microbiome) \sim treament + cohort  
$$

```{r}
# prepare input data
tmp.tse <- InputData.tse[,InputData.tse@colData$genotype != "WT"]
tmp.tse@colData$treatment <- relevel(tmp.tse@colData$treatment, "Vehicle")
tmp.tse@colData$treatment <- droplevels(tmp.tse@colData$treatment)
tmp.tse <- tmp.tse[rowSums(assay(tmp.tse, "metaphlan_RelAbund")) > 0,]

# define experiment name and output directories
maaslin3_expmt <- "treatments_vs_vehicle_in_ASO"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, maaslin3_expmt)

# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)

# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
treatments_vs_vehicle_in_ASO.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ treatment + cohort, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE, 
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

treatments_vs_vehicle_in_ASO.maaslin3_annotated <- list(
      abundance = treatments_vs_vehicle_in_ASO.maaslin3$fit_data_abundance$results %>% filter(metadata == "treatment"),
      prevalence = treatments_vs_vehicle_in_ASO.maaslin3$fit_data_prevalence$results %>% filter(metadata == "treatment")
    ) %>% 
      lapply(function(x) {
    taxonomy_info <- as.data.frame(rowData(tmp.tse))
    taxonomy_info_subset <- taxonomy_info[,1:which(colnames(taxonomy_info) == params$tax_level)]
    colnames(taxonomy_info_subset)[ncol(taxonomy_info_subset)] <- "feature"
    
    return(
      dplyr::full_join(taxonomy_info_subset, x, by = "feature") %>%
        arrange(qval_individual))}
    )

# save output tsvs as produced by maaslin3
system(paste0("cp ", maaslin3_tmp_outdir, "/*.tsv", " ", file.path(params$basic_outdir, params$tax_level, maaslin3_expmt)))
# save figures folder
system(paste0("cp -r ", file.path(output_directories.list$maaslin_tmpdir, maaslin3_expmt, "/figures"), " ", file.path(params$basic_outdir, params$tax_level, maaslin3_expmt)))

```

## Differentially prevalent {.tabset}

### Fp

```{r}
treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence %>% 
  filter(qval_individual < 0.1, value == "Fp") %>% 
  select(contains(colnames(treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence)), -contains(c("metadata", "value", "name", "null_"))) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```


### benCom

```{r}
treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence %>% 
  filter(qval_individual < 0.1, value == "benCom") %>% 
  select(contains(colnames(treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence)), -contains(c("metadata", "value", "name", "null_"))) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```


### Both

```{r}
dplyr::inner_join(
  treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence %>% 
  filter(qval_individual < 0.1, value == "Fp"),
  treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence %>% 
  filter(qval_individual < 0.1, value ==" benCom") %>% select(feature:N_not_zero),
  by = "feature", suffix = c(".Fp", ".benCom")) %>% 
  select(contains(colnames(treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence)), -contains(c("metadata", "value", "name", "null_"))) %>% 
  
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```

## Differentially abundant {.tabset}

### Fp

```{r}
treatments_vs_vehicle_in_ASO.maaslin3_annotated$abundance %>% 
  filter(qval_individual < 0.1, value == "Fp") %>% 
  select(contains(colnames(treatments_vs_vehicle_in_ASO.maaslin3_annotated$prevalence)), -contains(c("metadata", "value", "name", "null_"))) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```


### benCom

```{r}
treatments_vs_vehicle_in_ASO.maaslin3_annotated$abundance %>% 
  filter(qval_individual < 0.1, value == "benCom") %>% 
  select(contains(colnames(treatments_vs_vehicle_in_ASO.maaslin3_annotated$abundance)), -contains(c("metadata", "value", "name", "null_"))) %>% 
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```

### Both

```{r}
dplyr::inner_join(
  treatments_vs_vehicle_in_ASO.maaslin3_annotated$abundance %>% 
  filter(qval_individual < 0.1, value == "Fp"),
  treatments_vs_vehicle_in_ASO.maaslin3_annotated$abundance %>% 
  filter(qval_individual < 0.1, value ==" benCom") %>% select(feature:N_not_zero),
  by = "feature", suffix = c(".Fp", ".benCom")) %>% 
  select(contains(colnames(treatments_vs_vehicle_in_ASO.maaslin3_annotated$abundance)), -contains(c("metadata", "value", "name", "null_"))) %>% 
  
  DT::datatable(
    extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
    )
```

## Conclusions

  * No differentially **prevalent** features with Q-value < 0.1 comparing `Fp` w/
  `vehicle` controls
  
  * No differentially **prevalent** features with Q-value < 0.1 comparing `benCom` w/
  `vehicle` controls
  
  * Only 1 SGB statistically diff. **abundant** between Fp and vehicle, 4 between 
  benCom and vehicle
  
# Motor performances in response to treatments (Need data)
  
# Overall Conclusions

  * With the formula $Microbiome \sim genotype + cohort$ we do not see 
  statistically associated bacteria
  
  * Nor having only genotype as exposure (Not shown)
  
  * WT and ASO mice in this experiment do not show statistically significant 
  microbiome differences (N = 16 vs 16)
  
  * No statistical difference betweeen WT and ASO control mice 
  
  * No differentially **prevalent** features with Q-value < 0.1 comparing `Fp` w/
  `vehicle` controls
  
  * No differentially **prevalent** features with Q-value < 0.1 comparing `benCom` w/
  `vehicle` controls
  
  * Only 1 SGB statistically diff. **abundant** between Fp and vehicle, 4 between 
  benCom and vehicle

# Session Info

```{r}
sessionInfo()
```

  