---
title: "Probiotics to the rescue: F. prausnitzii and beneficial communities to re-establish motor performance in Parkinson mouse models"
author: "Anastasia Moiseyenko, Giacomo Antonello, Nicola Segata, Levi Waldron, Sarkis Mazmanian"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
params:
  basic_outdir: "results"
  microbiome_basic_transform: "RelAbund"
  tax_level: "SGB"
  beta_dist: "bray"
  beta_MDS: "MDS"
---

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
# packages for tabular data manipulation
library(tidyverse)
library(data.table)

# microbiome specific packages
library(maaslin3)
library(mia)
# some packages for data visualization (plots and tables)
library(miaViz)
library(pheatmap)
library(UpSetR)
library(ggpubr)
library(kableExtra)
library(reactable)

# MAC's package to download data
library(parkinsonsMetagenomicData)

library(ggh4x) # for complex ggplot2 faceting
# set ggplot2 plot themes 
theme_set(theme_light())
# set seed for reproducibility
set.seed(1234)
```

# Create output directories all in one chunk

```{r}

output_directories.list <- list(
  # alpha diversity directory
  alpha_div_outdir = file.path(params$basic_outdir, params$tax_level, "01_alpha_diversity"),
  # beta diversity output directory
  beta_div_outdir = file.path(params$basic_outdir, params$tax_level, "02_beta_diversity"),
  # differential abundance temporary directory
  maaslin_tmpdir = "~/Documents/maaslin3_tmp_results/mazmanian_matheus_proj4",
  # differential abundance output directory
  maaslin_outdir = file.path(params$basic_outdir, params$tax_level, "03_diff_abund"),
  # motor performance regression
  motorPerformance_outdir = file.path(params$basic_outdir, params$tax_level, "03_diff_abund", "motorPerformance")
)

tmp_out <- capture.output(sapply(output_directories.list, dir.create, showWarnings = FALSE, recursive = TRUE))
rm(tmp_out)
```

# Prepare data for analysis (run only once)

```{r, eval=interactive()}

# Load metadata
metadata.df <- read_tsv("Data/metadata_4.0.tsv") %>% 
  # format a few values in the metadata
  mutate(
    genotype = factor(genotype, levels = c("WT", "ASO")),
    cage = as.factor(cage),
    cohort = as.factor(cohort),
    subcohort = as.factor(subcohort),
    treatment = as.factor(treatment) %>% relevel(ref = "Vehicle")
    )

googleCloudStorageR::gcs_auth(json_file = "~/Downloads/curatedmetagenomicdata-232f4a306d1d.json")

# METAPHLAN4
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/get_reads_processed.R")
mpa_files_cache <- cacheMetagenomicData(metadata.df$uuid, data_type = "relative_abundance", ask_on_update = FALSE) %>% 
  # retrieve reads per sample too
  mutate(n_reads_processed = get_reads_processed.vec(cache_path))

# metaphlan <- get_metaphlan_locators(metadata.df$uuid, data_type = "relative_abundance")
metaphlan_profiles <- loadMetagenomicData(mpa_files_cache)

# the order of samples is preserved this way
metadata.df <- as.data.frame(dplyr::left_join(metadata.df, select(mpa_files_cache, uuid = UUID, n_reads_processed), by = "uuid"))
rownames(metadata.df) <- metadata.df$uuid
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/wrangle_metaphlan.R")
source("https://raw.githubusercontent.com/g-antonello/biobakeryUtils/refs/heads/main/R/complete_unknown_taxonomy.R")

mpa_raw <- cbind(
  as.data.frame(rowData(metaphlan_profiles)) %>%
    rownames_to_column("clade_name"), assay(metaphlan_profiles)) %>% 
  as_tibble() %>% 
  select(-additional_species)

dir.create("Data/metaphlan_profiles", recursive = TRUE, showWarnings = FALSE)
write_tsv(mpa_raw, "Data/metaphlan_profiles/metaphlan4_data.tsv")


# wrangle metaphlan4 data
mpa4_wrangled_taxonomy_mpa <- assay(metaphlan_profiles) %>%
  as.data.frame() %>% 
  rownames_to_column("clade_name") %>% 
  wrangle_metaphlan(taxonomic_lvl = params$tax_level)

InputData.tse <- TreeSummarizedExperiment(
  rowData = mpa4_wrangled_taxonomy_mpa$taxonomies,
  colData = metadata.df,
  assays = list(
    metaphlan_RelAbund = as.matrix(mpa4_wrangled_taxonomy_mpa$profiles/100)[,rownames(metadata.df)],
    metaphlan_Percent = as.matrix(mpa4_wrangled_taxonomy_mpa$profiles)[,rownames(metadata.df)]
    )
)
saveRDS(InputData.tse, "Data/InputData.tse.Rds")

```

```{r}
InputData.tse <- readRDS("Data/InputData.tse.Rds")
InputData.tse
```

# Dataset-wide microbiome composition statistics

## Number of taxa per taxonomic level

```{r}
InputData_splitByRanks.tse <- append(splitByRanks(InputData.tse), list("SGB" = InputData.tse))
 
InputData_splitByRanks.tse %>%  
  sapply(nrow)
```

## Prevalence distribution of each taxon

```{r}
InputData_splitByRanks.tse %>% 
  lapply(assay) %>% 
  lapply(apply, 1, function(x) sum(x > 0)) %>% 
  sapply(summary)
```

## Relative abundance of 30 most abundant bacteria per taxonomic level {.tabset}

```{r}
tmp <- InputData_splitByRanks.tse %>% 
  lapply(function(tse){
    max_n <- nrow(tse)
    top30MostAbund <- getTop(tse, assay.type = "metaphlan_RelAbund", method = "mean", top = min(c(30, max_n)))
    tse.subset <- tse[top30MostAbund,]
    return(tse.subset)
  })
  
```

```{r, results = "asis"}

for (tx in names(tmp)) {
  # Create header with cat()
  cat("\n\n###", tx, "\n")
  
  # Create plot and convert to plotly
  p <- miaViz::plotAbundance(tmp[[tx]], 
                            assay.type = "metaphlan_RelAbund",
                            group = tx) +
       facet_grid(cols = vars(cohort, treatment), scales = "free_x") + 
    theme(legend.position = "bottom")
  
  # Explicitly print plotly object
  print(p)
  
  cat("\n")
}

```

This is again to show that there are too many taxa to easily visualize them.
We need to rely on other methods.

## Heatmap all taxa and samples

```{r, fig.cap="Heatmap of full microbiome composition of the dataset. Relative abundances have been Z-scaled by taxon, to account for intrinsic differences in relative abundances in the microbiome"}

pheatmap::pheatmap(assay(InputData.tse) %>% magrittr::set_colnames(colData(InputData.tse)$sample_name), scale = "row", annotation_col = colData(InputData.tse) %>% as.data.frame() %>%  magrittr::set_rownames(.$sample_name) %>% select(treatment, cohort), angle_col = 315, show_rownames = FALSE)

```

# Analyses were be performed at the `r params$tax_level` level

Below the `TreeSummarizedExperiment` object summary that is going to be used
in all following analyses

```{r}
InputData.tse <- InputData_splitByRanks.tse[[params$tax_level]]

InputData.tse
```

