---
title: "Probiotics to the rescue: F. prausnitzii and beneficial communities to re-establish motor performance in Parkinson mouse models"
subtitle: "Metaphlan 4.1.1 relative abundances"
author: "Anastasia Moiseyenko, Giacomo Antonello, Nicola Segata, Levi Waldron, Sarkis Mazmanian"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    number_sections: true
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
params:
  basic_outdir: "results"
  microbiome_basic_transform: "RelAbund"
  tax_level: "SGB"
  beta_dist: "bray"
  beta_MDS: "MDS"
---

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
# packages for tabular data manipulation
library(tidyverse)
library(data.table)

# microbiome specific packages
library(maaslin3)
library(mia)
# some packages for data visualization (plots and tables)
library(miaViz)
library(pheatmap)
library(UpSetR)
library(ggpubr)
library(kableExtra)
library(reactable)

# MAC's package to download data
library(parkinsonsMetagenomicData)

library(ggh4x) # for complex ggplot2 faceting
# set ggplot2 plot themes 
theme_set(theme_light())
# set seed for reproducibility
set.seed(1234)

library(biobakeryUtils)

```

# Create output directories all in one chunk

```{r}

output_directories.list <- list(
  # alpha diversity directory
  alpha_div_outdir = file.path(params$basic_outdir, params$tax_level, "01_alpha_diversity"),
  # beta diversity output directory
  beta_div_outdir = file.path(params$basic_outdir, params$tax_level, "02_beta_diversity"),
  # differential abundance temporary directory
  maaslin_tmpdir = "~/Documents/maaslin3_tmp_results/mazmanian_Ana"
)

tmp_out <- capture.output(sapply(output_directories.list, dir.create, showWarnings = FALSE, recursive = TRUE))
rm(tmp_out)
```

# Prepare data for analysis

This is run only if interactive to avoid careless overwriting

```{r, eval=interactive()}

# Load metadata
metadata.df <- read_tsv("Data/metadata_4.0.tsv") %>% 
  # format a few values in the metadata
  mutate(
    genotype = as.factor(genotype) %>% relevel(ref = "WT"),
    cage = as.factor(cage),
    cohort = as.factor(cohort),
    subcohort = as.factor(subcohort),
    treatment = as.factor(treatment) %>% relevel(ref = "Vehicle")
    )

googleCloudStorageR::gcs_auth(json_file = "~/Downloads/curatedmetagenomicdata-232f4a306d1d.json")

# METAPHLAN4
cache_table <- cacheMetagenomicData(metadata.df$uuid, data_type = "relative_abundance", redownload = "no") 

# metaphlan <- get_metaphlan_locators(metadata.df$uuid, data_type = "relative_abundance")

InputMetaphlanData.tse <- suppressMessages(biobakeryUtils::loadMetagenomicData(cache_table))

# bypass the metadata including only what is important for us:

## get uuid and number_reads
important_pieces_from_parkinsonsMetagenomicData <- InputMetaphlanData.tse %>% 
  colData() %>% 
  as.data.frame() %>% 
  select(uuid, number_reads)

## generate a new colData data.frame with rows already ordered
new_colData <- as.data.frame(dplyr::left_join(metadata.df, important_pieces_from_parkinsonsMetagenomicData, by = "uuid"))
rownames(new_colData) <- new_colData$uuid
new_colData <- new_colData[colnames(InputMetaphlanData.tse),]
## replace the old colData with the new one
colData(InputMetaphlanData.tse) <- DataFrame(new_colData)

# create assay of Relative abundance as opposed to metaphlan's default percent
assay(InputMetaphlanData.tse, "metaphlan_RelAbund") <- assay(InputMetaphlanData.tse)/100
# order relative abundance [0,1]assay first 
assays(InputMetaphlanData.tse) <- rev(assays(InputMetaphlanData.tse))
# create a new assay with approximated counts as relAbund * reads_processed
assay(InputMetaphlanData.tse, "metaphlan_Counts") <- t(round(apply(assay(InputMetaphlanData.tse, "metaphlan_RelAbund"), 1, function(x) x * InputMetaphlanData.tse@colData$number_reads), 0))

# add phylogenetic tree
InputMetaphlanData.tse <-  AddPhyloTree_to_mpa_tse(data.tse = InputMetaphlanData.tse, CHOCOPhlAn_version = "202403")

# rename strain column into SGB
colnames(rowData(InputMetaphlanData.tse))[which(colnames(rowData(InputMetaphlanData.tse)) == "strain")] <- "SGB" 

saveRDS(InputMetaphlanData.tse, "Data/InputMetaphlanData.tse.Rds")

```

```{r}
InputMetaphlanData.tse <- readRDS("Data/InputMetaphlanData.tse.Rds")
InputMetaphlanData.tse
```

# Dataset-wide microbiome composition statistics

## Number of taxa per taxonomic level

```{r}
InputData_splitByRanks.tse <- append(splitByRanks(InputMetaphlanData.tse), list("SGB" = InputMetaphlanData.tse))
 
InputData_splitByRanks.tse %>%  
  sapply(nrow)
```

## Prevalence distribution of taxa per taxonomic rank

```{r}
InputData_splitByRanks.tse %>% 
  lapply(assay) %>% 
  lapply(apply, 1, function(x) sum(x > 0)) %>% 
  sapply(summary) %>% 
  kbl() %>% 
  kable_styling()
```

```{r}
n_taxa <- 20
```

## Relative abundance of `r n_taxa` most abundant bacteria per taxonomic level {.tabset}

```{r}

tmp <- InputData_splitByRanks.tse %>% 
  lapply(function(tse){
    max_n <- nrow(tse)
    topNMostAbund <- getTop(tse, assay.type = "metaphlan_RelAbund", method = "mean", top = min(c(n_taxa, max_n)))
    tse.subset <- tse[topNMostAbund,]
    return(tse.subset)
  })
  
```

```{r, results = "asis", fig.height=7, fig.width=10}

for (tx in names(tmp)) {
  # Create header with cat()
  cat("\n\n###", tx, "\n")
  
  # Create plot and convert to plotly
  p <- miaViz::plotAbundance(tmp[[tx]], 
                            assay.type = "metaphlan_RelAbund",
                            group = tx) +
       facet_nested(cols = vars(cohort, treatment), scales = "free_x") + 
    theme(legend.position = "bottom") + 
    guides(fill = guide_legend(nrow = 5))
  
  # Explicitly print plotly object
  print(p)
  
  cat("\n")
}
checkTaxonomy(tmp[[tx]])
```

This is again to show that there are too many taxa to easily visualize them.
We need to rely on other methods.

## Heatmap all taxa and samples

```{r, fig.cap="Heatmap of full microbiome composition of the dataset. Relative abundances have been Z-scaled by taxon, to account for intrinsic differences in relative abundances in the microbiome"}

pheatmap::pheatmap(assay(InputMetaphlanData.tse) %>% magrittr::set_colnames(colData(InputMetaphlanData.tse)$sample_name), scale = "row", annotation_col = colData(InputMetaphlanData.tse) %>% as.data.frame() %>%  magrittr::set_rownames(.$sample_name) %>% select(treatment, cohort), angle_col = 315, show_rownames = FALSE)

```

# Analyses are performed at the `r params$tax_level` level

Below the `TreeSummarizedExperiment` object summary that is going to be used
in all following analyses

```{r}
InputMetaphlanData.tse <- InputData_splitByRanks.tse[[params$tax_level]]

InputMetaphlanData.tse
```

# Alpha diversity does not vary in relation to sequencing depth

```{r}
InputMetaphlanData.tse <- addAlpha(InputMetaphlanData.tse, assay.type = "metaphlan_RelAbund", index = c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"))

lapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), 
       function(div) {
         stats_basic <- shapiro.test(as.data.frame(colData(InputMetaphlanData.tse))[[div]])
         text <- paste0("Shapiro-Wilk test: W = ", round(stats_basic$statistic, 2), "; P-value = ", round(stats_basic$p.value, 3))
         plot <- ggplot(as.data.frame(colData(InputMetaphlanData.tse)), aes(x = !!sym(div))) +
           geom_histogram() + 
           labs(
             title = div, 
             caption = text
           )
         
         }
       ) %>% 
  ggarrange(plotlist = .)

```

```{r, tab.cap="Assessment of dependence of alpha diversity measures in relation to sequencing depth"}
tmp <- sapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), function(div) {
  lm(as.formula(paste(div, "~", "number_reads")), data = colData(InputMetaphlanData.tse))
  
}, USE.NAMES = TRUE, simplify = FALSE)

table_results.df <- tmp %>% lapply(broom::tidy) %>% 
  bind_rows(.id = "AlphaIndex") %>% 
  filter(!grepl("Intercept", term))

table_results.df %>%
  kbl() %>% 
  row_spec(which(table_results.df$p.value < 0.05), bold = TRUE) %>% 
  kable_styling()
```

# Alpha diversity does not vary in relation to genotype, treatment and cohort

$$
AlphaIndex \sim genotype + treatment + cohort + \text{number_reads}
$$

```{r}
tmp <- sapply(c("dbp_dominance", "observed_richness", "shannon_diversity", "gini"), function(div) {
  lm(as.formula(paste(div, "~", "genotype + treatment + cohort + number_reads")), data = colData(InputMetaphlanData.tse))
  
}, USE.NAMES = TRUE, simplify = FALSE)

table_results.df <- tmp %>% lapply(broom::tidy) %>% 
  bind_rows(.id = "AlphaIndex") %>% 
  filter(!grepl("Intercept", term)) %>% 
  arrange(term) %>% 
  relocate(term)

write_tsv(table_results.df, file = file.path(output_directories.list$alpha_div_outdir, "alphas_vs_genotype_cohort_treatment.tsv"))

table_results.df %>%
  kbl() %>% 
  row_spec(which(table_results.df$p.value < 0.05), bold = TRUE, color = "red3") %>% 
  kable_styling()
```


# Variables and potential confounders

Variables:

  * treatment: the probiotic community given, or the empty liquid vehicle gavaged
  * genotype: mice genotype, either WT or ASO

Possible confounders:
  
  * Cohort: important to consider because it could determine inter-testing variability
  * Cage: Potential confounder, but mice receiving the same treatment are also in the same cage and there are only 2 mice per cage, making statics weak
  * Subcohort: Same as cage. It looks like a few samples were processed at a different time: 7W-F, 8W-F, 8W-B, 8C-B

# Beta diversity in relation to treatment and genotype

```{r}
dimRedName <- paste(params$beta_dist, "mds", sep = "_")
# this will work once mia or TreeSummarizedExperiment fixes some things
InputMetaphlanData.tse <- mia::addMDS(
  InputMetaphlanData.tse,
  assay.type = "metaphlan_RelAbund",
  method = params$beta_dist,
  ncomponents = 3, name = dimRedName
  )

variances_explained <- attr(reducedDim(InputMetaphlanData.tse, dimRedName), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

project1_meta_with_MDS.df <- cbind.data.frame(as.data.frame(colData(InputMetaphlanData.tse)), reducedDim(InputMetaphlanData.tse, dimRedName) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

beta_plot_basic <- project1_meta_with_MDS.df %>% 
ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = paste(genotype, treatment))) + 
  geom_point(size = 2) + 
  stat_ellipse(show.legend = FALSE) + 
  scale_color_manual(values = ggsci::pal_jco()(6)) +
  theme(legend.position = "top") +
  labs(color = "Genotype and Treatment",
       x = paste0("MDS 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("MDS 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )

```

```{r}
ggsave(plot = beta_plot_basic, filename = file.path(output_directories.list$beta_div_outdir, "bray_MDS_plot.png"), dpi = 600, height = 6.5, width = 7)

beta_plot_basic
```

## PERMANOVA statistics

```{r}

InputMetaphlanData.tse <- addPERMANOVA(
  InputMetaphlanData.tse,
  assay.type = "metaphlan_RelAbund",
  method = "bray",
  formula = x ~ genotype + treatment + cohort,
  name = "permanova",
  permutations = 999,
  test.homogeneity = TRUE
)

permanova_table.df <- PERMANOVA_to_table(InputMetaphlanData.tse)

write_tsv(permanova_table.df, file = file.path(output_directories.list$beta_div_outdir, paste(params$beta_dist, "adonisPermanova", "genotype", "treatment", "cohort.tsv", sep = "_")))
```

```{r}
permanova_table.df %>% 
  kbl() %>% 
  kable_styling()
```

# Microbiome differences between genotypes in treatment control mice

```
Are ASO mice having different microbiomes than WT, regardless of the experimental
cohort?
```

For this, I will include both cohorts together and adjust for cohort in the model.
Also, I will compare vehicle controls using WT as baseline.

$$
\log_2(Microbiome) ~ genotype + cohort
$$

```{r}
# prepare input data

tmp.tse <- InputMetaphlanData.tse[,InputMetaphlanData.tse@colData$treatment == "Vehicle"] 
tmp.tse@colData$genotype <- relevel(tmp.tse@colData$genotype, "WT") # not needed, but better safe than sorry
tmp.tse@colData$genotype <- droplevels(tmp.tse@colData$genotype)

tmp.tse

# define experiment name and output directories
maaslin3_expmt <- "genotype_vs_WT_vehicleCtrl_bothCohorts"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, maaslin3_expmt)
# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)
# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
set.seed(1234)
genotype_differences_in_ctrls.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ genotype + cohort, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE,
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

# add taxonomy
genotype_differences_in_ctrls.maaslin3_curated <- add_taxonomy_to_maaslin3(maaslin3_res = genotype_differences_in_ctrls.maaslin3, tmp.tse)
# write tables
write_maaslin3_curated_tables(genotype_differences_in_ctrls.maaslin3_curated, out.dir = maaslin3_final_outdir)
# save raw output
saveRDS(genotype_differences_in_ctrls.maaslin3, file.path(maaslin3_final_outdir, "maaslin3_raw_output.rds"))

```

## Diff. Prevalence - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- genotype_differences_in_ctrls.maaslin3_curated$prevalence %>% 
  filter(value == "ASO") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

## Barplots top 20 most diff. prevalent

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(cohort, genotype)), scales = "free_x") 
```

## Diff. Abundance - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- genotype_differences_in_ctrls.maaslin3_curated$abundance %>% 
  filter(value == "ASO") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```


## Barplot top 20 most diff. abundant

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(cohort, genotype)), scales = "free_x")
```

## Conclusions

There are no significantly affected microbes in ASO mice compared to WT.

What puzzles me is that there are many taxa that are uniquely identified by each
project leader (Matheus and Ana). Also, Matheus' sample size is 10 vs 10, Ana's 
is 8 vs 8 per cohort, in total 16 vs 16. We should not be less powerful than his.

What could be the factor here?

  * facility
  
  * handlers
  
  * gavage not sterile or affecting somehow?

  * diet
  
  * Could 2 weeks make such a striking difference in the composition?

# Cohort Fp7

```{r}
cohort_subdir <- "cohort_Fp7"

dir.create(file.path(params$basic_outdir, params$tax_level, cohort_subdir), showWarnings = FALSE, recursive = TRUE)

InputMetaphlanData_Fp7.tse <- InputMetaphlanData.tse[, InputMetaphlanData.tse@colData$cohort == "Fp7"] 
```

## Overall community trends (Beta diversity)

```{r}
dimRedName <- paste("Fp7", params$beta_dist, "mds", sep = "_")
# this will work once mia or TreeSummarizedExperiment fixes some things
InputMetaphlanData_Fp7.tse <- mia::addMDS(
  InputMetaphlanData_Fp7.tse,
  assay.type = "metaphlan_RelAbund",
  method = params$beta_dist,
  ncomponents = 3, name = dimRedName
  )

variances_explained <- attr(reducedDim(InputMetaphlanData_Fp7.tse, dimRedName), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

project1_meta_with_MDS.df <- cbind.data.frame(as.data.frame(colData(InputMetaphlanData_Fp7.tse)), reducedDim(InputMetaphlanData_Fp7.tse, dimRedName) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

beta_plot_basic <- project1_meta_with_MDS.df %>% 
ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = paste(genotype, treatment))) + 
  geom_point(size = 2) + 
  stat_ellipse(show.legend = FALSE) + 
  scale_color_manual(values = ggsci::pal_jco()(5)[c(3, 1, 2, 4, 5)]) +
  theme(legend.position = "top") +
  labs(color = "Genotype and Treatment",
       x = paste0("MDS 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("MDS 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )

```

```{r, fig.height=6, fig.width=6}
ggsave(plot = beta_plot_basic, filename = file.path(params$basic_outdir, params$tax_level, cohort_subdir, paste(params$beta_dist, params$beta_MDS, "plot.png", sep = "_")), dpi = 600, height = 6, width = 6)

beta_plot_basic
```

Same plot as above, without WT 

```{r}
dimRedName <- paste("Fp7", params$beta_dist, "mds", sep = "_")
# this will work once mia or TreeSummarizedExperiment fixes some things
InputMetaphlanData_Fp7_noWT.tse <- InputMetaphlanData_Fp7.tse[,InputMetaphlanData_Fp7.tse@colData$genotype != "WT"]

InputMetaphlanData_Fp7_noWT.tse <- mia::addMDS(
  InputMetaphlanData_Fp7_noWT.tse,
  assay.type = "metaphlan_RelAbund",
  method = params$beta_dist,
  ncomponents = 3, name = dimRedName
  )

variances_explained <- attr(reducedDim(InputMetaphlanData_Fp7_noWT.tse, dimRedName), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

project1_meta_with_MDS.df <- cbind.data.frame(as.data.frame(colData(InputMetaphlanData_Fp7_noWT.tse)), reducedDim(InputMetaphlanData_Fp7_noWT.tse, dimRedName) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

beta_plot_basic <- project1_meta_with_MDS.df %>% 
ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = paste(genotype, treatment))) + 
  geom_point(size = 2) + 
  stat_ellipse(show.legend = FALSE) + 
  scale_color_manual(values = ggsci::pal_jco()(5)[c(3, 1, 2, 4, 5)]) +
  theme(legend.position = "top") +
  labs(color = "Genotype and Treatment",
       x = paste0("MDS 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("MDS 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )

```

```{r, 6}
ggsave(plot = beta_plot_basic, filename = file.path(params$basic_outdir, params$tax_level, cohort_subdir, paste(params$beta_dist, params$beta_MDS, "noWT", "plot.png", sep = "_")), dpi = 600, height = 6, width = 6)

beta_plot_basic
```

```{r}

InputMetaphlanData_Fp7.tse <- addPERMANOVA(
  InputMetaphlanData_Fp7.tse,
  assay.type = "metaphlan_RelAbund",
  method = "bray",
  formula = x ~ genotype + treatment,
  name = "permanova",
  permutations = 999,
  test.homogeneity = TRUE
)

permanova_table.df <- PERMANOVA_to_table(InputMetaphlanData_Fp7.tse)

write_tsv(permanova_table.df, file = file.path(params$basic_outdir, params$tax_level, cohort_subdir, paste(params$beta_dist, "PERMANOVA.tsv", sep = "_")))
```

```{r}
permanova_table.df %>% 
  kbl() %>% 
  kable_styling()
```

## Pairwise group comparison (PERMANOVA and boxplot of Bray-Curtis distances)

```{r}
dissimName <- paste("Fp7", params$beta_dist, sep = "_")
InputMetaphlanData_Fp7.tse <- addDissimilarity(InputMetaphlanData_Fp7.tse, method = "bray", name = dissimName, assay.type= "metaphlan_RelAbund")
mtx <- metadata(InputMetaphlanData_Fp7.tse)[[dissimName]]
mtx[upper.tri(mtx,diag = TRUE)] <- NA

pairwise_dissim_bray <- reshape2::melt(mtx) %>% 
  filter(complete.cases(.)) %>%
  dplyr::inner_join(., as.data.frame(colData(InputMetaphlanData_Fp7.tse)) %>% transmute(Var1 = uuid, genotype_treatment = paste(genotype, treatment)), by = "Var1") %>% 
  dplyr::inner_join(., as.data.frame(colData(InputMetaphlanData_Fp7.tse)) %>% transmute(Var2 = uuid, genotype_treatment = paste(genotype, treatment)), by = "Var2") %>% 
  select(Var1, Var2, contains(c("uuid", "genotype", "treatment")), value) %>% 
  mutate(pairs = paste(genotype_treatment.y, genotype_treatment.x, sep = " vs "))

```

## Pairwise PERMANOVA

```{r}
source("https://raw.githubusercontent.com/pmartinezarbizu/pairwiseAdonis/refs/heads/master/pairwiseAdonis/R/pairwise.adonis.R")
set.seed(1234)
library(vegan)
pairwise_adonis_stats <- pairwise.adonis(x = as.dist(metadata(InputMetaphlanData_Fp7.tse)[[dissimName]]), 
                factors = paste(InputMetaphlanData_Fp7.tse@colData$genotype,InputMetaphlanData_Fp7.tse@colData$treatment), 
                perm = 9999
                )

kbl(pairwise_adonis_stats, caption = "Pairwise PERMANOVA test comparing groups as seen above") %>% 
  kable_styling()

```

```{r}
pairwise_dissim_bray %>% 
  filter(genotype_treatment.x != genotype_treatment.y) %>% 
ggboxplot(
  x = "pairs",
  y = "value") + 
  geom_point(position = position_jitter(width = 0.2, seed = 1234)) + 
  geom_text(aes(x = pairs, y = max(pairwise_dissim_bray$value+0.1), label = paste("Pairwise adonis2\nP-value =", p.value)), data = pairwise_adonis_stats) +
  
  labs(
    x = "",
    y = dissimName,
  )
```

## What microbes differ between `genotypes` in vehicle controls?

These bacteria should not be considered in the main exposure of interest later, 
as they vary because of genotype as well. Analyses should be performed
within each cohort.

Model formula is:

$$
\log_2(Microbiome) \sim genotype
$$

```{r}
# prepare input data

tmp.tse <- InputMetaphlanData_Fp7.tse[,InputMetaphlanData_Fp7.tse@colData$treatment == "Vehicle"] 
tmp.tse@colData$genotype <- relevel(tmp.tse@colData$genotype, "WT")
tmp.tse@colData$genotype <- droplevels(tmp.tse@colData$genotype)

tmp.tse

# define experiment name and output directories
maaslin3_expmt <- "genotype_vs_WT_in_treatment_vehicle"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir,  cohort_subdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, cohort_subdir, maaslin3_expmt)
# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)
# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
set.seed(1234)
genotype_differences_in_ctrls.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ genotype, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE,
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

# add taxonomy
genotype_differences_in_ctrls.maaslin3_curated <- add_taxonomy_to_maaslin3(maaslin3_res = genotype_differences_in_ctrls.maaslin3, tmp.tse)
# write tables
write_maaslin3_curated_tables(genotype_differences_in_ctrls.maaslin3_curated, out.dir = maaslin3_final_outdir)
# save raw output
saveRDS(genotype_differences_in_ctrls.maaslin3, file.path(maaslin3_final_outdir, "maaslin3_raw_output.rds"))
```

### Diff. Prevalence - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- genotype_differences_in_ctrls.maaslin3_curated$prevalence %>% 
  filter(value == "ASO") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplots top 20 most diff. prevalent

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x") 
```

### Diff. Abundance - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- genotype_differences_in_ctrls.maaslin3_curated$abundance %>% 
  filter(value == "ASO") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplot top 20 most diff. abundant

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x")
```

### Conclusions

  * There are no differentially prevalent bacteria comparing ASO vs WT 
  vehicle mice
  
  * 1 bacterium are decreased and 10 are increased in ASO mice.
  
  * The large effect sizes, however, point to a lack of statistical power to 
  confidently say that there are differences. This does not mean that there are 
  0 compositional shifts
  
  * Compared to Matheus' project 1 experiment, we find much fewer associations.
  Matheus' comparison was 10 vs 10, this comparison has 8 vs 8

## Effects of `treatment` in ASO male mice

Model formula is:

$$
\log_2(Microbiome) \sim treament  
$$

The analysis is restricted to ASO mice, and the baseline is ASO-vehicle mice.
This is allowed because we are interested in the impact of the communities on 
mice microbiome and their motor performance, and WT mice was not gavaged either
probiotic formulation, so they are incomparable.

```{r}
# prepare input data
tmp.tse <- InputMetaphlanData_Fp7.tse[,InputMetaphlanData_Fp7.tse@colData$genotype != "WT"]
tmp.tse@colData$treatment <- relevel(tmp.tse@colData$treatment, "Vehicle")
tmp.tse@colData$treatment <- droplevels(tmp.tse@colData$treatment)
tmp.tse <- tmp.tse[rowSums(assay(tmp.tse, "metaphlan_RelAbund")) > 0,]

tmp.tse

# define experiment name and output directories
maaslin3_expmt <- "treatments_vs_vehicle_in_ASO"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir, cohort_subdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, cohort_subdir, maaslin3_expmt)

# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)

# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
set.seed(1234)
treatments_vs_vehicle_in_ASO.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ treatment, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE, 
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

# add taxonomy
treatments_vs_vehicle_in_ASO.maaslin3_curated <- add_taxonomy_to_maaslin3(maaslin3_res = treatments_vs_vehicle_in_ASO.maaslin3, tmp.tse)
# write tables
write_maaslin3_curated_tables(treatments_vs_vehicle_in_ASO.maaslin3_curated, out.dir = maaslin3_final_outdir)
# save raw output
saveRDS(treatments_vs_vehicle_in_ASO.maaslin3, file.path(maaslin3_final_outdir, "maaslin3_raw_output.rds"))
```

### Diff. Prevalence - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- treatments_vs_vehicle_in_ASO.maaslin3_curated$prevalence %>% 
  filter(value == "Fp") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplots top 20 most diff. prevalent

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x") 
```

### Diff. Abundance - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- treatments_vs_vehicle_in_ASO.maaslin3_curated$abundance %>% 
  filter(value == "Fp") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplot top 20 most diff. abundant

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x")
```

### Conclusions

  * No differentially **prevalent** features with Q-value < 0.1 comparing `Fp`
  against `vehicle` ASO controls
  
  * No differentially **abundant** features with Q-value < 0.1 comparing `Fp`
  against `vehicle` ASO controls

  * Looking at raw data, I tend to conclude the same thing: a signal is there 
  but we are underpowered to label it "statistically significant".
  
# Cohort bCL

```{r}
cohort_subdir <- "cohort_bCL"

dir.create(file.path(params$basic_outdir, params$tax_level, cohort_subdir), showWarnings = FALSE, recursive = TRUE)

InputMetaphlanData_bCL.tse <- InputMetaphlanData.tse[, InputMetaphlanData.tse@colData$cohort == "bCL"] 
```

## Overall community trends (Beta diversity)

```{r}
dimRedName <- paste(params$beta_dist, "mds", sep = "_")
# this will work once mia or TreeSummarizedExperiment fixes some things
InputMetaphlanData_bCL.tse <- mia::addMDS(
  InputMetaphlanData_bCL.tse,
  assay.type = "metaphlan_RelAbund",
  method = params$beta_dist,
  ncomponents = 3, name = dimRedName
  )

variances_explained <- attr(reducedDim(InputMetaphlanData_bCL.tse, dimRedName), "eig")
prop_variances_explained <- variances_explained/sum(variances_explained)

project1_meta_with_MDS.df <- cbind.data.frame(as.data.frame(colData(InputMetaphlanData_bCL.tse)), reducedDim(InputMetaphlanData_bCL.tse, dimRedName) %>% as.data.frame() %>% set_names(paste("MDS", 1:ncol(.))))

beta_plot_basic <- project1_meta_with_MDS.df %>% 
ggplot(aes(x = `MDS 1`, y = `MDS 2`, color = paste(genotype, treatment))) + 
  geom_point(size = 2) + 
  stat_ellipse(show.legend = FALSE) + 
  scale_color_manual(values = ggsci::pal_jco()(5)[c(3, 1, 2, 4, 5)]) +
  theme(legend.position = "top") +
  labs(color = "Genotype and Treatment",
       x = paste0("MDS 1 [", round(prop_variances_explained[1]*100, 1), "%]"),
       y = paste0("MDS 2 [", round(prop_variances_explained[2]*100, 2), "%]")
  )

```

```{r, fig.height=6, fig.width=6}
ggsave(plot = beta_plot_basic, filename = file.path(params$basic_outdir, params$tax_level, cohort_subdir, paste(params$beta_dist, params$beta_MDS, "plot.png", sep = "_")), dpi = 600, height = 6, width = 6)

beta_plot_basic
```

```{r}

InputMetaphlanData_bCL.tse <- addPERMANOVA(
  InputMetaphlanData_bCL.tse,
  assay.type = "metaphlan_RelAbund",
  method = "bray",
  formula = x ~ genotype + treatment,
  name = "permanova",
  permutations = 999,
  test.homogeneity = TRUE
)

permanova_table.df <- PERMANOVA_to_table(InputMetaphlanData_bCL.tse)

write_tsv(permanova_table.df, file = file.path(params$basic_outdir, params$tax_level, cohort_subdir, paste(params$beta_dist, "PERMANOVA.tsv", sep = "_")))
```

```{r}
permanova_table.df %>% 
  kbl() %>% 
  kable_styling()
```

## Pairwise group comparison (PERMANOVA and boxplot of Bray-Curtis distances)

```{r}
dissimName <- paste("Fp7", params$beta_dist, sep = "_")
InputMetaphlanData_Fp7.tse <- addDissimilarity(InputMetaphlanData_Fp7.tse, method = "bray", name = dissimName, assay.type= "metaphlan_RelAbund")
mtx <- metadata(InputMetaphlanData_Fp7.tse)[[dissimName]]
mtx[upper.tri(mtx,diag = TRUE)] <- NA

pairwise_dissim_bray <- reshape2::melt(mtx) %>% 
  filter(complete.cases(.)) %>%
  dplyr::inner_join(., as.data.frame(colData(InputMetaphlanData_Fp7.tse)) %>% transmute(Var1 = uuid, genotype_treatment = paste(genotype, treatment)), by = "Var1") %>% 
  dplyr::inner_join(., as.data.frame(colData(InputMetaphlanData_Fp7.tse)) %>% transmute(Var2 = uuid, genotype_treatment = paste(genotype, treatment)), by = "Var2") %>% 
  select(Var1, Var2, contains(c("uuid", "genotype", "treatment")), value) %>% 
  mutate(pairs = paste(genotype_treatment.y, genotype_treatment.x, sep = " vs "))

```

## Pairwise PERMANOVA

```{r}
source("https://raw.githubusercontent.com/pmartinezarbizu/pairwiseAdonis/refs/heads/master/pairwiseAdonis/R/pairwise.adonis.R")
set.seed(1234)
library(vegan)
pairwise_adonis_stats <- pairwise.adonis(x = as.dist(metadata(InputMetaphlanData_Fp7.tse)[[dissimName]]), 
                factors = paste(InputMetaphlanData_Fp7.tse@colData$genotype,InputMetaphlanData_Fp7.tse@colData$treatment), 
                perm = 9999
                )

kbl(pairwise_adonis_stats, caption = "Pairwise PERMANOVA test comparing groups as seen above") %>% 
  kable_styling()

```

```{r}
pairwise_dissim_bray %>% 
  filter(genotype_treatment.x != genotype_treatment.y) %>% 
ggboxplot(
  x = "pairs",
  y = "value") + 
  geom_point(position = position_jitter(width = 0.2, seed = 1234)) + 
  geom_text(aes(x = pairs, y = max(pairwise_dissim_bray$value+0.1), label = paste("Pairwise adonis2\nP-value =", p.value)), data = pairwise_adonis_stats) +
  
  labs(
    x = "",
    y = dissimName,
  )
```

## What microbes differ between `genotypes` in vehicle controls?

These bacteria should not be considered in the main exposure of interest later, 
as they vary because of genotype as well. Analyses should be performed
within each cohort.

Model formula is:

$$
\log_2(Microbiome) \sim genotype
$$

```{r}
# prepare input data

tmp.tse <- InputMetaphlanData_bCL.tse[,InputMetaphlanData_bCL.tse@colData$treatment == "Vehicle"] 
tmp.tse@colData$genotype <- relevel(tmp.tse@colData$genotype, "WT")
tmp.tse@colData$genotype <- droplevels(tmp.tse@colData$genotype)

tmp.tse

# define experiment name and output directories
maaslin3_expmt <- "genotype_vs_WT_in_treatment_vehicle"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir,  cohort_subdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, cohort_subdir, maaslin3_expmt)
# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)
# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
set.seed(1234)
genotype_differences_in_ctrls.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ genotype, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE, 
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

# add taxonomy
genotype_differences_in_ctrls.maaslin3_curated <- add_taxonomy_to_maaslin3(maaslin3_res = genotype_differences_in_ctrls.maaslin3, tmp.tse)
# write tables
write_maaslin3_curated_tables(genotype_differences_in_ctrls.maaslin3_curated, out.dir = maaslin3_final_outdir)
# save raw output
saveRDS(genotype_differences_in_ctrls.maaslin3, file.path(maaslin3_final_outdir, "maaslin3_raw_output.rds"))
```

### Diff. Prevalence - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- genotype_differences_in_ctrls.maaslin3_curated$prevalence %>% 
  filter(value == "ASO") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplots top 20 most diff. prevalent

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x") 
```

### Diff. Abundance - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- genotype_differences_in_ctrls.maaslin3_curated$abundance %>% 
  filter(value == "ASO") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplot top 20 most diff. abundant

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x")
```

### Conclusions

  * No statistical differences due to genotype in bCL cohort comparing mice
  treated with `vehicle`.

## Effects of `treatment` in ASO male mice

Model formula is:

$$
\log_2(Microbiome) \sim treament  
$$

The analysis is restricted to ASO mice, and the baseline is ASO-vehicle mice.
This is allowed because we are interested in the impact of the communities on 
mice microbiome and their motor performance, and WT mice was not gavaged either
probiotic formulation, so they are incomparable.

```{r}
# prepare input data
tmp.tse <- InputMetaphlanData_bCL.tse[,InputMetaphlanData_bCL.tse@colData$genotype != "WT"]
tmp.tse@colData$treatment <- relevel(tmp.tse@colData$treatment, "Vehicle")
tmp.tse@colData$treatment <- droplevels(tmp.tse@colData$treatment)
tmp.tse <- tmp.tse[rowSums(assay(tmp.tse, "metaphlan_RelAbund")) > 0,]

# define experiment name and output directories
maaslin3_expmt <- "treatments_vs_vehicle_in_ASO"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir, cohort_subdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, cohort_subdir, maaslin3_expmt)

# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)

# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
set.seed(1234)
treatments_vs_vehicle_in_ASO.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ treatment, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE, 
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

# add taxonomy
treatments_vs_vehicle_in_ASO.maaslin3_curated <- add_taxonomy_to_maaslin3(maaslin3_res = treatments_vs_vehicle_in_ASO.maaslin3, tmp.tse)
# write tables
write_maaslin3_curated_tables(treatments_vs_vehicle_in_ASO.maaslin3_curated, out.dir = maaslin3_final_outdir)
# save raw output
saveRDS(treatments_vs_vehicle_in_ASO.maaslin3, file.path(maaslin3_final_outdir, "maaslin3_raw_output.rds"))
```

### Diff. Prevalence - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- treatments_vs_vehicle_in_ASO.maaslin3_curated$prevalence %>% 
  filter(value == "benCom") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplots top 20 most diff. prevalent

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x") 
```

### Diff. Abundance - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- treatments_vs_vehicle_in_ASO.maaslin3_curated$abundance %>% 
  filter(value == "benCom") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplot top 20 most diff. abundant

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(paste(treatment, genotype)), scales = "free_x")
```

### Conclusions

  * No statistically significant associations again, and again effect sizes 
  are very large both positively and negetively.

# Track probiotic species across all samples

## Fp (F. prausnitzii)

```{r}
Fp_species <- "s__Faecalibacterium_prausnitzii"
```


```{r}
tmp.tse <- InputMetaphlanData.tse
rownames(tmp.tse) <- collapse_taxonomy(as.data.frame(rowData(tmp.tse)))

tmp.tse %>% 
  meltSE(assay.type = "metaphlan_RelAbund", add.col = c("cohort", "genotype", "treatment")) %>%
  filter(grepl(paste(Fp_species, collapse = "|"), FeatureID)) %>% 
  group_by(FeatureID, genotype, cohort, treatment) %>% 
  reframe(
    mean_RelAbund = mean(metaphlan_RelAbund)
  ) %>% 
  DT::datatable(
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
  
```

## bCL (benCom community)

```{r}
benCom_species <- c("s__Roseburia_intestinalis", "s__Roseburia_faecis", "s__Fusicatenibacter_saccharivorans", "s__Bacteroides_vatus", "s__Eubacterium_rectale" , "s__Anaerostipes_hadrus", "s__Faecalibacterium_prausnitzii", "Prevotella_histicola")

benCom_species
```


```{r}
tmp.tse <- InputMetaphlanData.tse
rownames(tmp.tse) <- collapse_taxonomy(as.data.frame(rowData(tmp.tse)))

tmp.tse %>% 
  meltSE(assay.type = "metaphlan_RelAbund", add.col = c("cohort", "genotype", "treatment")) %>%
  filter(grepl(paste(benCom_species, collapse = "|"), FeatureID)) %>% 
  group_by(FeatureID, genotype, cohort, treatment) %>% 
  reframe(
    mean_RelAbund = mean(metaphlan_RelAbund)
  ) %>% 
  DT::datatable(
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

## Conclusions

    * No species from the Fp or benCom community was found
    
    * Time between sampling and last gavage?
    
    * Diet low in fibers?
    
    * Phylogenetically close bacteria may have taken the niche or serve a 
    similar purpose (i.e. butyrate production)?
    

# Are there cohort-specific bacteria in the dataset?

This can be asked on all data adjusting for other effects **OR** Comparing 
WT mice between the two cases (cleaner but low statistical power). 
I decided to do the first. 

$$
\log_2(Microbiome) \sim genotype + treatment + cohort
$$

```{r}
# prepare input data

tmp.tse <- InputMetaphlanData.tse 

tmp.tse

# define experiment name and output directories
maaslin3_expmt <- "cohort_assoc_bacteria"
maaslin3_tmp_outdir <- file.path(output_directories.list$maaslin_tmpdir,  cohort_subdir, maaslin3_expmt)
maaslin3_final_outdir <- file.path(params$basic_outdir, params$tax_level, cohort_subdir, maaslin3_expmt)
# crate temporary output directory
dir.create(maaslin3_tmp_outdir, showWarnings = FALSE, recursive = TRUE)
# create definitive output directory
dir.create(maaslin3_final_outdir, showWarnings = FALSE, recursive = TRUE)
```

Results are stored in ``r maaslin3_final_outdir``

```{r}
set.seed(1234)
cohort_differences.maaslin3 <- maaslin3(
    input_data = tmp.tse,
    formula = ~ genotype + treatment + cohort, 
    transform = "LOG",
    output = maaslin3_tmp_outdir,
    # do not standardize continuous metadata variables in the model
    standardize = FALSE, 
    plot_associations = TRUE,
    verbosity = "ERROR"
  )

# add taxonomy
cohort_differences.maaslin3_curated <- add_taxonomy_to_maaslin3(maaslin3_res = cohort_differences.maaslin3, tmp.tse)
# write tables
write_maaslin3_curated_tables(cohort_differences.maaslin3_curated, out.dir = maaslin3_final_outdir)
# save raw output
saveRDS(cohort_differences.maaslin3, file.path(maaslin3_final_outdir, "maaslin3_raw_output.rds"))
```

### Diff. Prevalence - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- cohort_differences.maaslin3_curated$prevalence %>% 
  filter(value == "Fp7") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplots top 20 most diff. prevalent

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(cohort), scales = "free_x") 
```

### Diff. Abundance - Top 20 with lowest crude p-value and highest absolute $\beta$

```{r}
top20_tmp_out <- cohort_differences.maaslin3_curated$abundance %>% 
  filter(value == "Fp7") %>% 
  arrange(pval_individual, desc(abs(coef))) %>% 
  head(20)

DT::datatable(
    top20_tmp_out,
     extensions = "Buttons",
    options = list(dom = "Bfrtip",
                   buttons = c('copy', 'csv', 'excel', 'pdf'))
  )
```

### Barplot top 20 most diff. abundant

```{r, fig.height=6, fig.width=20}

tmpTop20.tse <- tmp.tse[top20_tmp_out$feature,]
rownames(tmpTop20.tse) <- collapse_taxonomy(as.data.frame(rowData(tmpTop20.tse)))

miaViz::plotAbundance(tmpTop20.tse, assay.type = "metaphlan_RelAbund", group = "SGB") + 
  facet_grid(cols = vars(cohort), scales = "free_x")
```

# PCoA plot with just ASO-C v ASO-Fp (no WT)

```{r}

```

# Microbiome-typical plot panel

```{r}

```


# Motor performances in response to treatments (Data needed)
  
# Overall Conclusions (to be done)

# Session Info

```{r}
sessionInfo()
```

  